---
title: "Check correlation tables"
author: "Ian Hussey"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r include=FALSE}

# formatting options
# set default chunk options
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

# Correlation matrix checking

```{r}

library(dplyr)
library(readr)
library(tibble)
library(knitr)
library(kableExtra)

#'Correlation Boundary
#'
#'Computes the minimum and maximum values for a correlation between two variables in a correlation matrix
#'@param corMat A correlation matrix
#'@param var1 A number corresponding to the position in the matrix of the first variable of interest
#'@param var2 A number corresponding to the position in the matrix of the second variable of interest
#'@param method "default" computes the bound based of using values of 0 and pi in the correlative angle matrix. "greek" uses a closed form equation to compute.
#'@return A vector in which the first element is the minimum correlation and the second element is the maximum correlation
#'@examples
#'x <- matrix(c(1,.5,.5,0,.5,1,.5,.3,.5,.5,1,-.1,0,.3,-.1,1), nrow = 4, ncol = 4)
#'x[4,3] <- NA
#'x[3,4] <- NA
#'boundary(x, 4, 3)
#'@export
boundary <- function(corMat, var1, var2, method = "default") #find correlation boundary by NUMPACHAROEN's logic on page 407.
{
	if(nrow(corMat) != ncol(corMat))
	{
		stop("Matrix must be square!")
	}
	if(nrow(corMat) <= 1)
	{
		stop("Dimension must be greater than 1!")
	}
	n <- nrow(corMat)
	if(identical(method, "default"))
	{
		if(var1 == var2) #if elements are equal, their bounded correlation is 1
			return(c(1, 1))
		if(var1 == ncol(corMat)) #if var1 is in the last column, just move var2 to second to last
			corMat <- corSwap(corMat, var2, ncol(corMat) - 1)
		if(var2 == ncol(corMat)) #if var2 is in the last column, just move var1 to second to last
			corMat <- corSwap(corMat, var1, ncol(corMat) - 1)
		if(var2 != ncol(corMat) && var1 != ncol(corMat)) #if var1 and var2 are not in the last and second to last column, move them there
		{
			corMat <- corSwap(corMat, var1, ncol(corMat))
			corMat <- corSwap(corMat, var2, ncol(corMat) - 1)
		}
	upper <- corToAng(corMat)
	lower <- corToAng(corMat)
	upper[ncol(corMat), ncol(corMat)-1] <- 0 #set C[n, n-1] to 0 for maximum correlation
	lower[ncol(corMat), ncol(corMat)-1] <- pi #set C[n, n-1] to pi for minimum correlation
	return(c(angleToCor(lower)[ncol(corMat), ncol(corMat)-1],angleToCor(upper)[ncol(corMat), ncol(corMat)-1])) #convert new correlative angle matrices (upper and lower above) to correlation matrices, and return the value of C[n, n-1] which is now minimized and maximized.
	}
	if(identical(method, "greek"))
	{
		if(var1 == var2) #if elements are equal, their bounded correlation is 1
			return(c(1, 1))
		if(var1 == ncol(corMat)) #if var1 is in the last column, just move var2 to second to last
			corMat <- corSwap(corMat, var2, ncol(corMat) - 1)
		if(var2 == ncol(corMat)) #if var2 is in the last column, just move var1 to second to last
			corMat <- corSwap(corMat, var1, ncol(corMat) - 1)
		if(var2 != ncol(corMat) && var1 != ncol(corMat)) #if var1 and var2 are not in the last and second to last column, move them there
		{
			corMat <- corSwap(corMat, var1, ncol(corMat))
			corMat <- corSwap(corMat, var2, ncol(corMat) - 1)
		}
		theta <- corToAng(corMat)
		corr_alpha <- 1
		corr_beta <- cos(theta[n, (n - 1)])
		corr_gamma <- 1
		for(k in 1:(n - 2))
		{
			corr_alpha <- corr_alpha * sin(theta[(n - 1), k])
			corr_beta <- corr_beta * sin(theta[n, k])
		}
		for(k in 1:(n - 1))
		{
			corr_gamma <- corr_gamma * sin(theta[n, k])
		}
		lower <- corMat[n, (n - 1)] + (corr_alpha * (-1 * sqrt(corr_beta^2 + corr_gamma^2) - corr_beta))
		upper <- corMat[n, (n - 1)] + (corr_alpha * (sqrt(corr_beta^2 + corr_gamma^2) - corr_beta))
		return(c(lower, upper))
	}
}

#' Swaps columns and rows i and j in a correlation matrix
#' @param corMat A square correlation matrix
#' @param i, j Column and row indices to swap
#' @return The correlation matrix with swapped columns and rows
corSwap <- function(corMat, i, j) {
  if (i != j) {
    # Swap columns
    corMat[, c(i, j)] <- corMat[, c(j, i)]
    # Swap rows using the same indices to ensure symmetry
    corMat[c(i, j), ] <- corMat[c(j, i), ]
  }
  return(corMat)
}

angleToCor <- function(angMat) {
  if (!isSquareMatrix(angMat)) stop("Matrix must be square!")
  
  b <- matrix(nrow = nrow(angMat), ncol = ncol(angMat))
  for (i in seq_len(nrow(angMat))) {
    b[i, ] <- sapply(seq_len(ncol(angMat)), function(j) bsolve(angMat, i, j))
  }
  return(b %*% t(b))
}

corImpute <- function(corMat, method = "average", interval_prob = NA, interval = NA) {
  if (!isSquareMatrix(corMat)) stop("Matrix must be square!")
  if (countMiss(corMat) > 1) stop("Only 1 missing correlation currently supported!")
  
  bounds <- matSolve(corMat)
  if (is.null(bounds)) return(corMat)  # No missing values found

  position <- which(is.na(corMat), arr.ind = TRUE)
  replacement <- switch(method,
                        "average" = mean(bounds),
                        "lbound" = bounds[1],
                        "rbound" = bounds[2],
                        "min" = ifelse(bounds[1]^2 > bounds[2]^2, bounds[2], bounds[1]),
                        "max" = ifelse(bounds[1]^2 > bounds[2]^2, bounds[1], bounds[2]),
                        "uniform" = runif(1, min = bounds[1], max = bounds[2]),
                        "custom" = customBound(bounds, interval_prob, interval),
                        stop("Invalid method specified!"))
  corMat[position] <- replacement
  return(corMat)
}

customBound <- function(bounds, interval_prob, interval) {
  if (any(is.na(interval_prob), is.na(interval), length(interval_prob)*2 != length(interval), sum(interval_prob) != 1)) {
    stop("Invalid custom interval specifications!")
  }
  interval_pick <- sample(seq_along(interval_prob), size = 1, prob = interval_prob)
  bound_range <- bounds[1] + diff(bounds) * interval[c(interval_pick * 2 - 1, interval_pick * 2)]
  runif(1, min = bound_range[1], max = bound_range[2])
}

isSquareMatrix <- function(mat) {
  nrow(mat) == ncol(mat) && nrow(mat) > 1
}

countMiss <- function(corMat) {
  sum(lower.tri(corMat, diag = TRUE) & is.na(corMat))
}

bsolve <- function(angMat, i, j) #implement NUMPACHAROEN algorithm
{
	if(nrow(angMat) != ncol(angMat))
	{
		stop("Matrix must be square!")
	}
	dimension = nrow(angMat)
	if(i == 1 && j == 1)
		return(1)
	if(i >= 2 && j == 1)
		return(cos(angMat[i,j]))
	if(i == j && i >= 2 && j <= dimension)
	{
		answer <- 1
		for(index in 1:(j - 1))
		{
			answer <- answer * sin(angMat[i,index])
		}
		return(answer)
	}
	if(j >= 2 && j <= (i - 1))
	{
		answer <- cos(angMat[i,j])
		for(index in 1:(j - 1))
		{
			answer <- answer * sin(angMat[i,index])
		}
		return(answer)
	}
	if(j >= (i + 1) && j <= dimension)
		return(0)
}

cholesky <- function(mat) #custom Cholesky-Crout algorithm to handle missing data
{
	x <- matrix(0, ncol = NCOL(mat), nrow = NROW(mat))
	for(j in 1:ncol(mat))
	{
		for(i in j:nrow(mat))
		{
			if(j == i)
			{
				temp = 0
				if((j - 1) != 0)
				{
					for(k in 1:(j - 1))
					{
						temp = temp + (x[j, k]^2)
					}
				}
				x[i, j] = sqrt(mat[j, j] - temp)
			}
			else
			{
				temp = 0
				if((j - 1) != 0)
				{
					for(k in 1:(j - 1))
					{
						temp = temp + (x[i, k] * x[j, k])
					}
				}
				x[i, j] = (mat[i, j] - temp)/x[j, j]
			}
		}
	}
	x[is.nan(x)] <- 0
	return(x)
}

thetaSolve <- function(bMat, i, j, partAng=NA) #solves for theta using logic on NUMPACHAROEN (2013) page 405, partial angle matrix can be provided to speed up computation (so code can access what it already has calculated previously instead of having to compute everything from the beginning column again)
{
  if(!is.matrix(partAng) || !is.data.frame(partAng))
    partAng <- matrix(NA, ncol = NCOL(bMat), nrow = NROW(bMat))
  if(nrow(bMat) != ncol(bMat))
  {
    stop("Matrix must be square!")
  }
  dimension = nrow(bMat)
  if(j >= i)
    return(0)
  if(i >= 2 && j == 1)
    return(acos(bMat[i, j]))
  if(j >= 2 && j <= (i - 1))
  {
    sinprod <- 1
    for(index in 1:(j - 1))
    {
      if(is.na(partAng[i, index]))
        sinprod <- sinprod * sin(thetaSolve(bMat, i, index))
      else
        sinprod <- sinprod * sin(partAng[i, index])
    }
    if(is.na(sinprod))
      return(NA)
    if(sinprod == 0)
      return(0)
    else
    {
      if(is.na((bMat[i, j]/sinprod)))
        return(NA)
      if((bMat[i, j]/sinprod) > 1) #If product is out of the domain of acos, treat it as acos(1)
        return(0)
      if((bMat[i, j]/sinprod) < -1) #If product is out of the domain of acos, treat it as acos(-1)
        return(pi)
      else
       return(acos(bMat[i, j]/sinprod))
    }
  }
}

#'Solves for the boundary of one missing correlation
#'
#'Automatically detects and returns the boundary of one missing correlation in a correlation matrix
#'@param corMat A correlation matrix with one missing correlation
#'@return A vector in which the first element is the minimum correlation and the second element is the maximum correlation
#'@examples
#'x <- matrix(c(1,.5,.5,0,.5,1,.5,.3,.5,.5,1,-.1,0,.3,-.1,1), nrow = 4, ncol = 4)
#'x[4,3] <- NA
#'x[3,4] <- NA
#'matSolve(x)
#'@export
matSolve <- function(corMat) #automatically finds the boundaries of one missing entry in a correlative matrix
{
	if(countMiss(corMat) > 1)
		stop("Correlation matrix has more than 1 missing entry!")
	for(i in 1:nrow(corMat))
	{
		for(j in 1:i)
		{
			if(is.na(corMat[i,j]))
			{
				return(boundary(corMat, i, j))
			}
		}
	}
}

triangle_to_cor_matrix <- function(triangle){
  
  library(tibble)
  
  # if input is a df or tibble, convert to matrix
  if(is.data.frame(triangle) | is_tibble(triangle)){
    triangle <- as.matrix(triangle)
  } 
  if(!is.matrix(triangle)){
    stop("Input must be a matrix, data frame, or tibble")
  }
  
  # assess if the triangle is an upper triangle. I.e., are all the lower tri values NA and all the upper tri values non-NA? If so, transpose the matrix to make it a lower triangle. 
  if(all(is.na(triangle[lower.tri(triangle)])) &
     all(!is.na(triangle[upper.tri(triangle)]))){
    triangle <- t(triangle)
  }
  
  # create an empty matrix of the same dimension as 'triangle' to hold the mirrored values
  mirror_matrix <- matrix(0, nrow = nrow(triangle), ncol = ncol(triangle))
  
  # assign the lower triangle of 'mirror_matrix' with the lower triangle of 'triangle'
  mirror_matrix[lower.tri(mirror_matrix)] <- triangle[lower.tri(triangle)]
  
  # add the transposed 'mirror_matrix' to itself to complete the mirroring process
  full_matrix <- mirror_matrix + t(mirror_matrix) - diag(nrow(triangle))
  
  # fill diagonal with 1s
  diag(full_matrix) <- 1
  
  return(full_matrix)
}

#' Correlation Matrix to Correlative Angle Matrix
#'
#' Transform a correlation matrix to its correlative angle matrix form
#' @param corMat A matrix which contains the correlations between variables
#' @return A correlative angle matrix corresponding to the input correlation matrix
#' @examples
#'x <- matrix(c(1,.5,.5,0,.5,1,.5,.3,.5,.5,1,-.1,0,.3,-.1,1), nrow = 4, ncol = 4)
#'x[4,3] <- NA
#'x[3,4] <- NA
#'y <- corToAng(x)
#'y[4,3] <- 0
#'round(angleToCor(y), 2)
#' @export
corToAng <- function(corMat) #function to loop over a correlation matrix calling thetaSolve on each element to get the theta values for the correlative angle matrix
{
	if(nrow(corMat) != ncol(corMat))
	{
		stop("Matrix must be square!")
	}
	if(nrow(corMat) <= 1)
	{
		stop("Dimension must be greater than 1!")
	}
	angMat <- matrix(NA, nrow = nrow(corMat), ncol = ncol(corMat))
	b <- cholesky(corMat)
	for(j in 1:ncol(angMat))
	{
		for(i in 1:nrow(angMat))
		{
			angMat[i,j] <- thetaSolve(b, i, j, angMat)
		}
	}
	return(angMat)
}

data_test <- read_csv("test_data_correct.csv", show_col_types = FALSE) |>
  triangle_to_cor_matrix()

data_test_missing <- data_test

data_test_missing[1,3] <- NA
data_test_missing[3,1] <- NA

data_test[1,3]
round(corImpute(data_test_missing, method = "average"), 2)[1,3]
round(corImpute(data_test_missing, method = "lbound"), 2)[1,3]
round(corImpute(data_test_missing, method = "rbound"), 2)[1,3]

```

### Impute individual correlations

```{r}

dat <- matrix(data = c())
dat[1,3] <- NA
dat[3,1] <- NA

dat[1,3]
round(corImpute(data_test_missing, method = "average"), 2)[1,3]
round(corImpute(data_test_missing, method = "lbound"), 2)[1,3]
round(corImpute(data_test_missing, method = "rbound"), 2)[1,3]

```

## Addition for error checking

```{r}

triangle_to_cor_matrix <- function(triangle){
  
  # if input is a df or tibble, convert to matrix
  if(is.data.frame(triangle) | is_tibble(triangle)){
    triangle <- as.matrix(triangle)
  } 
  if(!is.matrix(triangle)){
    stop("Input must be a matrix, data frame, or tibble")
  }
  
  # assess if the triangle is an upper triangle. I.e., are all the lower tri values NA and all the upper tri values non-NA? If so, transpose the matrix to make it a lower triangle. 
  if(all(is.na(triangle[lower.tri(triangle)])) &
     all(!is.na(triangle[upper.tri(triangle)]))){
    triangle <- t(triangle)
  }
  
  # create an empty matrix of the same dimension as 'triangle' to hold the mirrored values
  mirror_matrix <- matrix(0, nrow = nrow(triangle), ncol = ncol(triangle))
  
  # assign the lower triangle of 'mirror_matrix' with the lower triangle of 'triangle'
  mirror_matrix[lower.tri(mirror_matrix)] <- triangle[lower.tri(triangle)]
  
  # add the transposed 'mirror_matrix' to itself to complete the mirroring process
  full_matrix <- mirror_matrix + t(mirror_matrix) - diag(nrow(triangle))
  
  # fill diagonal with 1s
  diag(full_matrix) <- 1
  
  return(full_matrix)
}

positive_definite <- function(correlation_matrix){
  all(eigen(correlation_matrix)$values > 0)
}

within_bounds <- function(correlation_matrix){
  all(correlation_matrix <= +1) & all(correlation_matrix >= -1)
}

check_correlation_matrix <- function(data_matrix) {
  require(stats)  # Ensure the stats package is loaded for rank function
  
  original_matrix <- data_matrix
  imputed_matrix <- matrix(NA, nrow = nrow(data_matrix), ncol = ncol(data_matrix))
  diff_matrix <- matrix(NA, nrow = nrow(data_matrix), ncol = ncol(data_matrix))
  
  # Function to impute a single element
  singleImpute <- function(mat, row, col) {
    mat[row, col] <- NA  # Introduce NA at the specified position
    imputed_value <- corImpute(mat, method = "average")
    return(imputed_value[row, col])
  }
  
  # Perform imputation and calculate differences
  for (i in 1:nrow(data_matrix)) {
    for (j in 1:ncol(data_matrix)) {
      suppressWarnings({
        imputed_matrix[i, j] <- singleImpute(data_matrix, i, j)
        diff_matrix[i, j] <- original_matrix[i, j] - imputed_matrix[i, j]
      })
    }
  }
  
  # Format matrix with original and difference
  formatted_matrix <- matrix("", nrow = nrow(data_matrix), ncol = ncol(data_matrix))
  abs_diff <- abs(diff_matrix)
  ranks <- matrix(rank(-abs_diff, ties.method = "min"), nrow = nrow(data_matrix))  # Ensure ranks is a matrix

  for (i in 1:nrow(data_matrix)) {
    for (j in 1:ncol(data_matrix)) {
      stars <- ifelse(ranks[i, j] == 1, "***",
                      ifelse(ranks[i, j] == 2, "**",
                             ifelse(ranks[i, j] == 3, "*", "")))
      #formatted_matrix[i, j] <- sprintf("%.2f [%.2f]%s", janitor::round_half_up(original_matrix[i, j], 2), janitor::round_half_up(diff_matrix[i, j], 2), stars)
      formatted_matrix[i, j] <- sprintf("%.2f [%+.2f]%s", janitor::round_half_up(original_matrix[i, j], 2), -janitor::round_half_up(diff_matrix[i, j], 2), stars)
      #formatted_matrix[i, j] <- sprintf("%+.2f [%+.2f]%s", janitor::round_half_up(original_matrix[i, j], 2), janitor::round_half_up(diff_matrix[i, j], 2), stars)
    }
  }
  
  upper_tri_only <- function(mat){
    diag(mat) <- NA
    mat[upper.tri(mat)] <- NA
    t(mat)
  }
  
  original_matrix <- upper_tri_only(original_matrix)
  imputed_matrix <- upper_tri_only(imputed_matrix)
  diff_matrix <- upper_tri_only(diff_matrix)
  formatted_matrix <- upper_tri_only(formatted_matrix)
  
  return(list(
    original_matrix = original_matrix,
    original_is_pd = positive_definite(triangle_to_cor_matrix(original_matrix)),
    original_is_within_bounds = within_bounds(triangle_to_cor_matrix(original_matrix)),
    imputed_matrix = imputed_matrix,
    difference_matrix = diff_matrix,
    formatted_results = formatted_matrix
  ))
}

```

### Example

```{r}

# load data
data_df <- read_csv("test_data_correct.csv", show_col_types = FALSE) |>
  janitor::clean_names()

col_names <- colnames(data_df)

data_test <- data_df |>
  triangle_to_cor_matrix()

#data_test
#data_test[4,1]
data_test_temp <- data_test
data_test_temp[4, 1] <- 0.7
data_test_temp[1, 4] <- 0.7

```

```{r}

# fit results
res <- check_correlation_matrix(data_test_temp)


# print results
options(knitr.kable.NA = '')

# res$original_matrix |>
#   kable(caption = "Reported correlation matrix") |>
#   kable_classic(full_width = FALSE)
# 
# res$imputed_matrix |> 
#   janitor::round_half_up(2) |>
#   kable(caption = "Imputed correlation matrix") |>
#   kable_classic(full_width = FALSE)
# 
# res$difference_matrix |> 
#   janitor::round_half_up(2) |>
#   kable(caption = "Difference matrix") |>
#   kable_classic(full_width = FALSE)

tibble(test = c("All correlations within the bounds [-1, 1]",
                "Matrix is positive-definite"),
       result = c(res$original_is_within_bounds,
                  res$original_is_pd)) |>
  kable() |>
  kable_classic(full_width = FALSE)

res$formatted_results |>
  kable(caption = "Reported Correlation [Imputed Correlation Difference]") |>
  #kable() |>
  kable_classic(full_width = FALSE)

```


```{r}

# with column and row names
res_for_printing <- res$formatted_results
colnames(res_for_printing) <- col_names

bind_cols(
  data.frame("Var" = col_names),
  as.data.frame(res_for_printing)
) |>
  #kable(caption = "Results") |>
  kable() |>
  kable_classic(full_width = TRUE)

```

## Write to disk

```{r}

library(writexl)

df_original_matrix <- as.data.frame(res$original_matrix)
df_imputed_matrix <- as.data.frame(res$imputed_matrix)
df_difference_matrix <- as.data.frame(res$difference_matrix)
df_formatted_results <- as.data.frame(res$formatted_results)

df_tests <- data.frame(is_positive_definite = res$original_is_pd,
                       does_not_violate_bounds = res$original_is_within_bounds)

colnames(df_original_matrix) <- col_names
colnames(df_imputed_matrix) <- col_names
colnames(df_difference_matrix) <- col_names
colnames(df_formatted_results) <- col_names

write_xlsx(list(original = df_original_matrix,
                tests = df_tests,
                imputed = df_imputed_matrix,
                difference = df_difference_matrix,
                combined = df_formatted_results),
           "cor_check_results.xlsx")

```

# Gemini alternative

Gemini suggests that Frank's original code is a literal implementation of the math from the cited papers, but is computationally inefficient and may even have some bugs. It suggests this alternative:

```{r}

#' Efficient Correlation Boundary
#'
#' Computes the min/max values for a missing correlation using the 
#' closed-form algebraic method (Schur complement/Projection).
#'
#' @param corMat A correlation matrix.
#' @param var1 Index of the first variable whose correlation bounds should be estimated.
#' @param var2 Index of the second variable whose correlation bounds should be estimated.
#' @return A vector c(min, max).
#' @export
boundary_algebraic <- function(corMat, var1, var2) {
  
  # Helper to return consistent dataframe structure for edge cases
  return_df <- function(mean, lower, upper) {
    data.frame(imputed_r_mean = mean, 
               imputed_r_lower_bound = lower, 
               imputed_r_upper_bound = upper)
  }

  # 1. Validation
  if(nrow(corMat) != ncol(corMat)) stop("Matrix must be square!")
  
  # Edge Case: Same variable
  if(var1 == var2) return(return_df(1, 1, 1))
  
  # 2. Identify the "Known" variables (K)
  idx_k <- setdiff(seq_len(nrow(corMat)), c(var1, var2))
  
  # Edge Case: No conditioning variables (2x2 matrix) -> Bounds are -1 to 1
  if(length(idx_k) == 0) return(return_df(0, -1, 1))
  
  # 3. Extract submatrices
  r_1k <- corMat[var1, idx_k]
  r_2k <- corMat[var2, idx_k]
  R_kk <- corMat[idx_k, idx_k]
  
  # 4. Solve using Linear Algebra
  R_kk_inv <- tryCatch(solve(R_kk), error = function(e) NULL)
  
  # Edge Case: Singular matrix
  if(is.null(R_kk_inv)) {
    warning("Matrix K is singular; bounds might be loose or inaccurate.")
    return(return_df(0, -1, 1))
  }
  
  # 5. Calculation
  center <- as.numeric(t(r_1k) %*% R_kk_inv %*% r_2k)
  
  var1_resid <- max(0, 1 - as.numeric(t(r_1k) %*% R_kk_inv %*% r_1k))
  var2_resid <- max(0, 1 - as.numeric(t(r_2k) %*% R_kk_inv %*% r_2k))
  
  radius <- sqrt(var1_resid * var2_resid)
  
  # Success Return
  return(return_df(center, center - radius, center + radius))
}

```

usage

```{r}

data(mtcars)

# Create a correlation matrix from the first 4 variables:
# 1:mpg, 2:cyl, 3:disp, 4:hp
M <- cor(mtcars[, 1:4])


v1_idx <- 1 # mpg
v2_idx <- 4 # hp

# Note: We pass the full matrix M. The function ignores M[1,4] and M[4,1] 
# and uses the other constraints to find what those values *could* be.
bounds <- boundary_algebraic(M, v1_idx, v2_idx)


# original
round(M[v1_idx, v2_idx], 3)
# bounds
bounds

```

## simulation 

```{r}
multiverse_plot <- function(data, 
                            outcome = "outcome", 
                            outcome_name = "Outcome",
                            interval_lower = NULL,
                            interval_upper = NULL,
                            rank_by = "rank",
                            rank_by_outcome = FALSE,
                            relative_height_of_upper_plot = 0.70, 
                            outcome_cutoff = NULL){
  
  # dependencies
  require(dplyr)
  require(tidyr)
  require(ggplot2)
  require(cowplot)
  require(ggtext)

  # TODO consider checking inputs with rlang::ensym(variable)

  # rename outcome variable
  data <- data |>
    rename(outcome = {{outcome}},
           rank = {{rank_by}})
  
  # if intervals aren't specified, set them to the same value as outcome (so they are invisible in the plot)
  if(!is.null(interval_lower)){
    data <- data |>
      rename(interval_lower = {{interval_lower}})
  } else {
    data$interval_lower <- data$outcome
  }
  
  if(!is.null(interval_upper)){
    data <- data |>
      rename(interval_upper = {{interval_upper}})
  } else {
    data$interval_upper <- data$outcome
  }
  
  # if rank_by_outcome, rank by the outcome. otherwise, take order that was passed.
  if(rank_by_outcome){
    data <- data |>
      arrange(outcome) |>
      mutate(rank = row_number())
  }
  
  # ensure all columns other than rank and outcome are factors
  data <- data |>
    mutate(across(.cols = -c(rank, outcome, interval_lower, interval_upper),
                  .fns = as.factor))
  
  p_estimates <- data |>
    mutate(outcome_name = outcome_name) |>
    ggplot(aes(rank, outcome)) +
    geom_linerange(aes(ymin = interval_lower, ymax = interval_upper)) +
    geom_point(shape = "circle", size = 2) +
    facet_grid(outcome_name ~ ., space = "free_y", scales = "free_y", switch = "y") +
    scale_y_continuous(breaks = scales::pretty_breaks()) + 
    scale_x_continuous(NULL, expand = c(.02, .02)) +
    ylab("Simulation results") +
    #theme_classic() +
    theme_minimal() +
    theme(legend.position = "none",
          axis.line.x = element_blank(),
          strip.placement = "outside",
          strip.background = element_rect(fill = NA, colour = NA),
          panel.spacing.x = unit(0.15, "cm"),
          strip.text.y = element_markdown(angle = 180, face = "bold", size = 7),
          axis.text.y = element_text(angle = 0, 
                                     hjust = 0.5, 
                                     size = 6),
          axis.title.y = element_text(size = 9, face = "bold"),
          panel.spacing = unit(0.25, "lines"),
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
  
  if(!is.null(outcome_cutoff)){
    p_estimates <- p_estimates + geom_hline(yintercept = outcome_cutoff, linetype = "dotted")
  }
  
  p_specs <- data |> 
    tidyr::pivot_longer(cols = c(-"rank", -"outcome", -"interval_lower", -"interval_upper")) |> 
    arrange(rank) |>
    ggplot(aes(x = rank, y = factor(value), color = name)) + 
    geom_point(size = 2, shape = "square") +
    facet_grid(name ~ ., space = "free_y", scales = "free_y", switch = "y") +
    guides(color = "none") +
    scale_x_continuous(NULL, expand = c(.02, .02)) +
    ylab("Simulation conditions") +
    #theme_classic() +
    theme_minimal() +
    theme(strip.placement = "outside",
          strip.background = element_rect(fill = NA, colour = NA),
          panel.spacing.x = unit(0.15, "cm"),
          strip.text.y = element_markdown(angle = 180, face = "bold", size = 7),
          axis.text.y = element_text(angle = 0, hjust = 1, size = 6),
          axis.title.y = element_text(size = 9, face = "bold"),
          panel.spacing = unit(0.25, "lines"),
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank()) +
    scale_color_brewer(palette = "Dark2")

  p_multiverse <- cowplot::plot_grid(p_estimates, 
                                     p_specs, 
                                     axis = "bltr", 
                                     align = "v", 
                                     ncol = 1, 
                                     rel_heights = c(relative_height_of_upper_plot, 1))
  
  return(p_multiverse)
}

```

### static matrix

```{r}

#' Generate Equicorrelation Matrix
#'
#' Creates an N x N matrix with 1s on the diagonal and a static value 'r' on the off-diagonals.
#'
#' @param n Integer. The dimension of the matrix (number of rows/columns).
#' @param r Numeric. The correlation value for off-diagonal elements.
#' @param check_validity Logical. If TRUE, checks if the resulting matrix is positive semi-definite.
#' @return An N x N matrix.
#' @examples
#' # Generate a valid 4x4 matrix with correlation 0.5
#' m <- generate_static_matrix(4, 0.5)
#' 
#' # Generate a matrix that is technically invalid (negative eigenvalues)
#' m_bad <- generate_static_matrix(3, -0.9)
#' @export
generate_static_matrix <- function(k, r, check_validity = TRUE) {
  
  if (k < 2) stop("Dimension k must be at least 2.")
  if (abs(r) > 1) warning("Correlation r should typically be between -1 and 1.")

  # 1. Create the matrix
  # Create an n*n matrix filled with 'r'
  mat <- matrix(r, nrow = k, ncol = k)
  
  # Set the diagonal to 1
  diag(mat) <- 1
  
  # 2. Check mathematical validity (Positive Semi-Definite)
  # For an equicorrelation matrix, r must satisfy:
  # -1 / (n - 1) <= r <= 1
  if (check_validity) {
    lower_bound <- -1 / (k - 1)
    if (r < lower_bound) {
      warning(sprintf(
        "Matrix is not positive semi-definite (invalid correlation matrix). 
         For dimension %d, r must be >= %.4f. You provided %.4f.",
        k, lower_bound, r
      ))
    }
  }
  
  return(mat)
}

generate_and_test <- function(k, r, var1, var2){
  generate_static_matrix(k = k, r = r) %>%
    boundary_algebraic(corMat = ., var1 = var1, var2 = var2) |> 
    mutate(bounds_width = imputed_r_upper_bound - imputed_r_lower_bound)
}



# res <- 
#   expand_grid(
#       k = 20,
#       r = .5,
#       var1 = 1,
#       var2 = seq.int(2, 20)
#     ) |>
#   mutate(res = purrr::pmap(.l = list(k = k, r = r, var1 = var1, var2 = var2),
#                            .f = generate_and_test)) |>
#   unnest(res) |>
#   mutate_if(is.numeric, roundwork::round_up, digits = 2)

res <-
  expand_grid(
    k = seq.int(2, 20),
    r = c(seq(0, .9, by = .1), .99) |> round_up(2),
    var1 = 1,
    var2 = 2
  ) |>
  mutate(res = purrr::pmap(.l = list(k = k, r = r, var1 = var1, var2 = var2),
                           .f = generate_and_test)) |>
  unnest(res) |>
  mutate_if(is.numeric, roundwork::round_up, digits = 2)

```

multiverse plot

```{r fig.height=9, fig.width=9}

data_summary_ordered <- res |>
  # the outcome variable must be called `outcome`
  # eg the estimand of your simulation, whether the false positive rate, estimated effect size, etc.
  mutate(outcome = bounds_width) |>
  
  # # order the rows within the columns in the way you want them to appear left to right
  # arrange(k, r, var1, var2) |>
  # mutate(rank = row_number()) |>

  # note that you might instead want to rank by the outcome variable rather than the conditions
  arrange(outcome) |>
  mutate(rank = row_number()) |>
  
  # select only the columns that you had multiple values for in the expand.grid() call in the simulation as well as `outcome` and `rank`
  # rename them if necessary to plot nicely
  dplyr::select(`r` = r, 
                `k size matrix` = k, 
                outcome,
                rank)

p_multiverse <- data_summary_ordered |>
  multiverse_plot(outcome_name = "r bounds")

p_multiverse

```


### uniform distributed matrix with varying bounds location

```{r}

#' Generate Random Uniform Correlation Matrix
#'
#' Creates an N x N matrix where off-diagonal elements are sampled from a 
#' uniform distribution. Returns NULL if the resulting matrix is not 
#' positive semi-definite.
#'
#' @param n Integer. The dimension of the matrix.
#' @param min_r Numeric. The minimum value for the uniform distribution.
#' @param max_r Numeric. The maximum value for the uniform distribution.
#' @return An N x N matrix if valid, or NULL if invalid.
#' @export
generate_uniform_matrix <- function(k, min_r = -1, max_r = 1) {
  
  if (k < 2) stop("Dimension n must be at least 2.")
  if (min_r > max_r) stop("min_r cannot be greater than max_r.")

  # 1. Initialize matrix with 1s on diagonal
  mat <- matrix(1, nrow = k, ncol = k)
  
  # 2. Generate random values for the upper triangle
  # Number of unique off-diagonal elements
  n_upper <- k * (k - 1) / 2 
  random_vals <- runif(n_upper, min = min_r, max = max_r)
  
  # 3. Fill upper triangle and mirror to lower to ensure symmetry
  mat[upper.tri(mat)] <- random_vals
  # Transpose the upper triangle to fill the lower triangle
  mat[lower.tri(mat)] <- t(mat)[lower.tri(mat)]
  
  # 4. Check for Positive Semi-Definiteness (PSD)
  # Calculate eigenvalues. If the smallest eigenvalue is negative, 
  # the matrix is not a valid correlation matrix.
  eigenvalues <- eigen(mat, symmetric = TRUE, only.values = TRUE)$values
  
  # Use a small tolerance (1e-8) for floating point comparisons
  if (min(eigenvalues) < -1e-8) {
    return(NULL)
  }
  
  return(mat)
}


generate_and_test <- function(k, min_r, max_r, var1, var2) {
  
  # 1. Attempt to generate matrix
  mat <- generate_uniform_matrix(k = k, min_r = min_r, max_r = max_r)
  
  # 2. STOP if invalid: If mat is NULL (not Positive Definite), return NULL immediately.
  # This prevents the NULL from hitting 'mutate'
  if (is.null(mat)) return(NULL)
  
  # 3. Calculate bounds on valid matrix
  # You can use 'possibly' here if you fear other errors, but strictly checking 
  # is.null(mat) solves the immediate crash.
  res <- boundary_algebraic(corMat = mat, var1 = var1, var2 = var2)
  
  # 4. Add width column
  res |> 
    mutate(bounds_width = imputed_r_upper_bound - imputed_r_lower_bound)
}

library(furrr)

plan(multisession, workers = future::availableCores())
  
set.seed(42)

res2 <-
  expand_grid(
    k = seq.int(3, 12),
    max_r = seq(0, .9, by = .1) |> round_up(2),
    var1 = 1,
    var2 = 2,
    iteration = 1:1000
  ) |>
  mutate(min_r = max_r -.3) |>
  mutate(res = furrr::future_pmap(.l = list(k = k, min_r = min_r, max_r = max_r, var1 = var1, var2 = var2),
                                  .f = generate_and_test,
                                  .progress = TRUE,
                                  .options = furrr_options(seed = TRUE))) |>
  unnest(res) 

res_summary2 <- res2 |>
  group_by(k,
           min_r,
           max_r,
           var1,
           var2) |>
  summarize(n_pd = n(),
            bounds_width_mean = mean(bounds_width),
            bounds_width_min = min(bounds_width),
            bounds_width_max = max(bounds_width)) |>
  ungroup() |>
  mutate_if(is.numeric, roundwork::round_up, digits = 2)

```

multiverse plot

```{r fig.height=9, fig.width=9}

data_summary_ordered2 <- res_summary2 |>
  # the outcome variable must be called `outcome`
  # eg the estimand of your simulation, whether the false positive rate, estimated effect size, etc.
  mutate(outcome = bounds_width_mean) |>
  mutate(r_pop_mean = (max_r + min_r)/2) |>
  
  # order the rows within the columns in the way you want them to appear left to right
  arrange(k, r_pop_mean) |>
  mutate(rank = row_number()) |>

  # # note that you might instead want to rank by the outcome variable rather than the conditions
  # arrange(outcome) |>
  # mutate(rank = row_number()) |>
  
  # select only the columns that you had multiple values for in the expand.grid() call in the simulation as well as `outcome` and `rank`
  # rename them if necessary to plot nicely
  
  dplyr::select(r_pop_mean,
                `k size matrix` = k, 
                outcome,
                bounds_width_min,
                bounds_width_max,
                rank)

p_multiverse2 <- data_summary_ordered2 |>
  multiverse_plot(outcome_name = "r bounds",
                  interval_lower = "bounds_width_min",
                  interval_upper = "bounds_width_max")

p_multiverse2

res_summary2 |>
  mutate(r_pop_mean = (max_r + min_r)/2) |>
  ggplot(aes(r_pop_mean, n_pd)) +
  geom_point()

```


### uniform distributed matrix with varying bounds width

```{r}

#' Generate Random Uniform Correlation Matrix
#'
#' Creates an N x N matrix where off-diagonal elements are sampled from a 
#' uniform distribution. Returns NULL if the resulting matrix is not 
#' positive semi-definite.
#'
#' @param n Integer. The dimension of the matrix.
#' @param min_r Numeric. The minimum value for the uniform distribution.
#' @param max_r Numeric. The maximum value for the uniform distribution.
#' @return An N x N matrix if valid, or NULL if invalid.
#' @export
generate_uniform_matrix <- function(k, min_r = -1, max_r = 1) {
  
  if (k < 2) stop("Dimension n must be at least 2.")
  if (min_r > max_r) stop("min_r cannot be greater than max_r.")

  # 1. Initialize matrix with 1s on diagonal
  mat <- matrix(1, nrow = k, ncol = k)
  
  # 2. Generate random values for the upper triangle
  # Number of unique off-diagonal elements
  n_upper <- k * (k - 1) / 2 
  random_vals <- runif(n_upper, min = min_r, max = max_r)
  
  # 3. Fill upper triangle and mirror to lower to ensure symmetry
  mat[upper.tri(mat)] <- random_vals
  # Transpose the upper triangle to fill the lower triangle
  mat[lower.tri(mat)] <- t(mat)[lower.tri(mat)]
  
  # 4. Check for Positive Semi-Definiteness (PSD)
  # Calculate eigenvalues. If the smallest eigenvalue is negative, 
  # the matrix is not a valid correlation matrix.
  eigenvalues <- eigen(mat, symmetric = TRUE, only.values = TRUE)$values
  
  # Use a small tolerance (1e-8) for floating point comparisons
  if (min(eigenvalues) < -1e-8) {
    return(NULL)
  }
  
  return(mat)
}


generate_and_test <- function(k, min_r, max_r, var1, var2) {
  
  # 1. Attempt to generate matrix
  mat <- generate_uniform_matrix(k = k, min_r = min_r, max_r = max_r)
  
  # 2. STOP if invalid: If mat is NULL (not Positive Definite), return NULL immediately.
  # This prevents the NULL from hitting 'mutate'
  if (is.null(mat)) return(NULL)
  
  # 3. Calculate bounds on valid matrix
  # You can use 'possibly' here if you fear other errors, but strictly checking 
  # is.null(mat) solves the immediate crash.
  res <- boundary_algebraic(corMat = mat, var1 = var1, var2 = var2)
  
  # 4. Add width column
  res |> 
    mutate(bounds_width = imputed_r_upper_bound - imputed_r_lower_bound)
}

library(furrr)

plan(multisession, workers = future::availableCores())
  
set.seed(42)

res <-
  expand_grid(
    k = seq.int(3, 12),
    max_r = seq(0, .9, by = .1) |> round_up(2),
    var1 = 1,
    var2 = 2,
    iteration = 1:1000
  ) |>
  mutate(min_r = max_r * -1) |>
  mutate(res = furrr::future_pmap(.l = list(k = k, min_r = min_r, max_r = max_r, var1 = var1, var2 = var2),
                                  .f = generate_and_test,
                                  .progress = TRUE,
                                  .options = furrr_options(seed = TRUE))) |>
  unnest(res) 

res_summary <- res |>
  group_by(k,
           min_r,
           max_r,
           var1,
           var2) |>
  summarize(n_pd = n(),
            bounds_width_mean = mean(bounds_width),
            bounds_width_min = min(bounds_width),
            bounds_width_max = max(bounds_width)) |>
  ungroup() |>
  mutate_if(is.numeric, roundwork::round_up, digits = 2)

```

multiverse plot

```{r fig.height=9, fig.width=9}

data_summary_ordered <- res_summary |>
  # the outcome variable must be called `outcome`
  # eg the estimand of your simulation, whether the false positive rate, estimated effect size, etc.
  mutate(outcome = bounds_width_mean) |>
  mutate(r_pop_range = max_r - min_r) |>
  
  # order the rows within the columns in the way you want them to appear left to right
  arrange(k, r_pop_range) |>
  mutate(rank = row_number()) |>

  # # note that you might instead want to rank by the outcome variable rather than the conditions
  # arrange(outcome) |>
  # mutate(rank = row_number()) |>
  
  # select only the columns that you had multiple values for in the expand.grid() call in the simulation as well as `outcome` and `rank`
  # rename them if necessary to plot nicely
  
  dplyr::select(r_pop_range,
                `k size matrix` = k, 
                outcome,
                bounds_width_min,
                bounds_width_max,
                rank)

p_multiverse <- data_summary_ordered |>
  multiverse_plot(outcome_name = "r bounds",
                  interval_lower = "bounds_width_min",
                  interval_upper = "bounds_width_max")

p_multiverse

```


# The hard limit of PDness for r and k

```{r fig.height=6, fig.width=6}

library(ggplot2)
library(dplyr)

# 1. Generate Data for the theoretical limit
# Range of dimensions k from 2 to 25
plot_data <- data.frame(k = seq(2, 25, by = 0.1)) |>
  mutate(min_r = -1 / (k - 1))

# 2. Create the Plot
ggplot(plot_data, aes(x = k, y = min_r)) +
  # Add the geometric limit line
  geom_line(size = 1.2, color = "black") +
  
  # Shade the "Impossible" region (Below the line)
  geom_ribbon(aes(ymin = -1, ymax = min_r), fill = "#d73027", alpha = 0.3) +
  
  # Shade the "Possible" region (Above the line, up to 0 for this view)
  geom_ribbon(aes(ymin = min_r, ymax = 0), fill = "#4575b4", alpha = 0.3) +
  
  # Add Labels to the regions
  annotate("text", x = 15, y = -0.6, label = "Impossible Region\n(Not Positive Definite)", 
           color = "#d73027", fontface = "bold", size = 5) +
  annotate("text", x = 6, y = -0.02, label = "Valid Region", 
           color = "#4575b4", fontface = "bold", size = 5, vjust = 1) +

  # Add specific points for context (The "Crowded Room" examples)
  geom_point(data = data.frame(k = c(2, 3, 5, 11), y = c(-1, -0.5, -0.25, -0.1)), 
             aes(x = k, y = y), size = 3) +
  geom_text(data = data.frame(k = c(2, 3, 5, 11), y = c(-1, -0.5, -0.25, -0.1)),
            aes(x = k, y = y, label = paste0("k=", k, "\nmin r=", y)), 
            nudge_y = 0.08, nudge_x = 0.5, size = 3.5, lineheight = 0.9) +

  # Formatting
  scale_y_continuous(limits = c(-1, 0), breaks = seq(-1, 0, 0.1), name = "Correlation (r)") +
  scale_x_continuous(breaks = seq(2, 24, 2), name = "Number of Variables (k)") +
  theme_minimal(base_size = 14) +
  labs(
    title = "The 'Crowded Room' Problem",
    subtitle = "Minimum possible uniform correlation as dimension (k) increases",
    caption = "Formula: min_r = -1 / (k - 1)"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

```

# Generating PD correlation matrices


Gram Matrix Method: Generate a random rectangular matrix $X$ and compute $C = \text{cov}(X)$. This is always PD by definition.

```{r}

# Simple way to get a random valid correlation matrix
generate_valid_random <- function(n) {
  # Generate random data
  X <- matrix(rnorm(n * n * 2), ncol = n) 
  # Return its correlation matrix
  return(cor(X))
}

generate_valid_random(10)

```

The "Onion" Method (Joe, 2006) - (formally described by Lewandowski, Kurowicka, and Joe, 2009, extending Joe, 2006) is the standard solution to the problem you just encountered.: Generates random correlations uniformly distributed over the valid elliptope.

The Algorithm
The method relies on the partial correlation property.
- Start with a correlation matrix $R_{k-1}$.
- To extend it to $R_k$, we need to generate a new column of correlations $r$.
- The condition for $R_k$ to be positive definite is that the quadratic form $r^T R_{k-1}^{-1} r < 1$.
- This inequality describes a hyper-ellipsoid. The method samples a direction vector (on a sphere) and a radius (beta distribution) to find a valid $r$.

R Implementation

Here is a standalone function that implements the Onion Method. This will never return NULL. It will always produce a valid correlation matrix.
You can tune alpha (the shape parameter of the Beta distribution):
- alphad = 1: Uniform distribution over the space of correlation matrices (rarely extreme correlations).
- alphad > 1: Concentrated around the identity matrix (correlations closer to 0).
- alphad < 1: More density near the edges (high correlations).

```{r}

#' Generate Random Correlation Matrix (Onion Method)
#' 
#' Generates a random correlation matrix that is guaranteed to be 
#' Positive Semi-Definite. Based on the method by Lewandowski, Kurowicka, and Joe (2009).
#'
#' @param d The dimension of the matrix.
#' @param eta The parameter controlling the distribution. 
#'            1 = Uniform distribution of matrices. 
#'            Higher values = Correlations concentrated near 0.
#' @return A valid d x d correlation matrix.
generate_onion_matrix <- function(d, eta = 1) {
  
  # 1. Initialize the first element (1x1 matrix)
  beta <- eta + (d - 2) / 2
  u <- rbeta(1, beta, beta)
  r12 <- 2 * u - 1
  R <- matrix(c(1, r12, r12, 1), nrow = 2) # Start with 2x2
  
  # 2. Iterate to add layers (from dimension 3 to d)
  if (d > 2) {
    for (k in 3:d) {
      
      # Sampling y from Beta distribution (radius in the ellipsoid)
      beta <- eta + (d - 1 - k) / 2
      y <- rbeta(1, k / 2 - 1, beta)
      r <- sqrt(y) # Radius
      
      # Sample direction uniformly from the unit sphere
      # (Normalizing a vector of standard normals)
      theta <- rnorm(k - 1)
      theta <- theta / sqrt(sum(theta^2))
      
      # The new column is formed by mapping the direction/radius 
      # back into the correlation space using the Cholesky root of the previous R.
      # L is the lower triangular Cholesky factor of the current R
      L <- t(chol(R))
      
      # The new correlations w
      w <- L %*% (r * theta)
      
      # Update the matrix
      # New R is:
      # [ Old R   w ]
      # [ w^T     1 ]
      R_new <- matrix(0, nrow = k, ncol = k)
      R_new[1:(k-1), 1:(k-1)] <- R
      R_new[k, 1:(k-1)] <- w
      R_new[1:(k-1), k] <- w
      R_new[k, k] <- 1
      
      R <- R_new
    }
  }
  
  return(R)
}

```