---
title: "Getting started with port"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with port}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r setup}
library(port)
```

## The question

A correlation matrix is only valid if it is **positive semidefinite** (PSD).
Reported matrices in papers, though, are *rounded* — you see `0.62`, not the
real number behind it. So the honest question is not "is this exact matrix PSD?"
but

> given that every off-diagonal was rounded to `d` decimals, does **any** PSD
> matrix exist inside the rounding box around the reported values?

`port` answers that **box** question. Writing `delta = 0.5 * 10^(-d)`, the box
is `X_ii = 1`, and for `i != j`, `X_ij` in
`[R_ij - delta, R_ij + delta]` intersected with `[-1, 1]`. The verdict is one of:

- **`inconsistent`** — no PSD matrix fits the box, so the reported matrix cannot be
  a genuine correlation matrix (a rigorous certificate; this is the payload);
- **`consistent`** — a PSD matrix in the box is exhibited (or none was ruled out);
- **`undecided`** — precision-limited; the box grazes the PSD boundary.

## Checking a matrix

A genuine correlation matrix, rounded, comes back `consistent`:

```{r}
R_ok <- matrix(c(1,    0.30, 0.20,
                 0.30, 1,    0.25,
                 0.20, 0.25, 1), 3, 3)
check_corr_psd(R_ok, decimals = 2)
```

An inconsistent one comes back `inconsistent`. The classic case is a "frustrated
triangle": variables 1 and 2 are both strongly correlated with 3, yet reported as
strongly *negatively* correlated with each other.

```{r}
R_bad <- matrix(c(1,    0.9,  0.9,
                  0.9,  1,   -0.9,
                  0.9, -0.9,  1), 3, 3)
res <- check_corr_psd(R_bad, decimals = 2)
res
```

### The certificate and the rigor guarantee

An `inconsistent` verdict carries a **witness vector** `v`: a direction along which
no in-box matrix has a nonnegative quadratic form, so none can be PSD. Extract it
with `certificate()`:

```{r}
certificate(res)
```

The guarantee is that the verdict is **floating-point-safe**: the witness
quantity `v'Rv + delta * (||v||_1^2 - ||v||_2^2)` is evaluated with a rigorous
a-priori error bound (Higham 2002), so `inconsistent` never rests on a solver
tolerance. A corollary is soundness in the other direction: because the bound is
a sound over-estimate, a box that contains any PSD matrix can never be reported
`inconsistent`. In particular a real correlation matrix rounded to `d` decimals is
never a false positive.

Every verdict also carries a `certified` flag. `inconsistent` is always certified;
a `consistent` is certified when an explicit in-box PSD matrix was exhibited and
Rump-verified, and marked *presumed* otherwise ("not shown inconsistent").

Be precise about what `inconsistent` proves: the reported values cannot be the
**complete-data, single-sample Pearson** correlation matrix of any dataset, up
to rounding. Genuinely non-PSD reported matrices can arise legitimately — from
pairwise deletion, polychoric/tetrachoric estimation, meta-analytically
assembled cells, or upstream disattenuation — so the print enumerates these
cautions, and the severity output (below) is the discriminator: those mechanisms
rarely produce violations far beyond the rounding step.

A useful headline statistic folds the verdict and the severity into one number
in precision units — the smallest reporting imprecision that could excuse the
matrix:

```{r}
excusable_delta(R_bad)
```

Here no conventional rounding could excuse the matrix: entries would have to be
mis-stated by more than ±`r round(excusable_delta(R_bad), 2)` each.

## Screening a corpus

`check_corr_psd_batch()` runs the check over a list of matrices and returns a
tidy table — useful for screening many reported matrices at once.

```{r}
mats <- list(ok = R_ok, frustrated_triangle = R_bad, identity = diag(4))
check_corr_psd_batch(mats, decimals = 2, quiet = TRUE)
```

## Localizing the fault

When a matrix is inconsistent, `localize_psd_fault()` asks **where** the problem is
and **how severe** it is. Non-PSDness is a global property, so a unique culprit
is often under-identified; the tool reports an honest localization class rather
than inventing one. The verdict is one of `cell`, `cell_tentative`, `triad`,
`variable`, `joint`, `diffuse`, or `none`.

### A frustrated triangle is a `triad`

Fixing *any one* of the three correlations would resolve the violation, so the
fault is not separable to a single cell:

```{r}
localize_psd_fault(R_bad, decimals = 2)
```

### A single bad cell

Here one transcription-style error makes a 5-variable matrix inconsistent. The tool
localizes it to the offending cell, states the interval that cell *should* have
occupied, and how far beyond rounding it must move:

```{r}
R_cell <- matrix(c( 1,    0.56,  0.68,  0.05,  0.82,
                    0.56, 1,    -0.26, -0.31, -0.75,
                    0.68, -0.26, 1,    -0.18,  0.50,
                    0.05, -0.31, -0.18, 1,     0.27,
                    0.82, -0.75, 0.50,  0.27,  1), 5, 5, byrow = TRUE)
localize_psd_fault(R_cell, decimals = 2)
```

### A joint fault (no single cell suffices)

Two independent inconsistencies mean no single edit repairs the matrix:

```{r}
blk <- matrix(c(1, 0.9, 0.9, 0.9, 1, -0.9, 0.9, -0.9, 1), 3, 3)
R_joint <- matrix(0, 6, 6); R_joint[1:3, 1:3] <- blk; R_joint[4:6, 4:6] <- blk
diag(R_joint) <- 1
localize_psd_fault(R_joint, decimals = 2)$localization_verdict
```

### The raw evidence

Every verdict carries the full box-aware evidence behind it — per-cell implied
intervals (A), rounding-robust inconsistent triples (B), leave-one-out (C), the
sparse minimal correction (D), and the R² / VIF variable localizer. Get it with
`fault_evidence()`:

```{r}
ev <- fault_evidence(localize_psd_fault(R_bad, decimals = 2))
names(ev)
```

The inconsistent-triple scan (component B) is also exposed on its own; it is pure
interval arithmetic and needs no solver:

```{r}
inconsistent_triples(R_bad, decimals = 2)
```

## Benign causes come first

Non-PSDness is *often benign*. A composite score plus its subscores, or a full
set of category dummies, is exactly rank-deficient by construction — not a sign
of a data problem (van Tilburg & van Tilburg 2023; Lorenzo-Seva & Ferrando
2021). Beyond structure, whole classes of *reported* matrices are legitimately
non-PSD: pairwise-deletion tables (each cell from different available cases),
polychoric/tetrachoric matrices, meta-analytically assembled cells, and
already-disattenuated matrices. For a forensic tool this matters, so every
inconsistent result carries a `structural` diagnosis that separates a **benign
near-boundary** rank-deficiency from a **substantive over-the-boundary**
violation, and the print lists the generators to rule out before inferring a
reporting error.

Here a composite equal to the sum of three subscores is nudged just past the
rounding boundary. The cellular verdict is `diffuse` (there is no single cell to
blame), but the structural read names the likely cause:

```{r}
Sig <- matrix(0.5, 3, 3); diag(Sig) <- 1
a <- c(1, 1, 1); r <- as.numeric(Sig %*% a) / sqrt(sum(a * (Sig %*% a)))
R_comp <- rbind(cbind(Sig, r), c(r, 1))
for (k in 1:3) R_comp[k, 4] <- R_comp[4, k] <- r[k] + 0.02   # nudge past the boundary
lf <- localize_psd_fault(round(R_comp, 2), decimals = 2)
lf$structural$severity_class
lf$structural$pattern
```

Contrast this with the frustrated triangle, whose severity is far beyond rounding:

```{r}
localize_psd_fault(R_bad, decimals = 2)$structural$severity_class
```

## Plausibility, even when consistent

When a matrix passes the inconsistency test it can still sit implausibly close to
the boundary. For `consistent` matrices the result reports a **plausibility
gradient**: the per-variable reported-point R² and its closest approach to the
100% ceiling. The exactly-singular composite matrix (before the nudge) rounds to
`consistent`, but one variable's R² exceeds 100% at face value:

```{r}
R_exact <- round(rbind(cbind(Sig, r), c(r, 1)), 2)   # composite = sum of subscores
lf0 <- localize_psd_fault(R_exact, decimals = 2)
lf0$localization_verdict        # "none": consistent given rounding
lf0$plausibility$max_r2         # ... but a variable's reported-point R^2 exceeds 1
```

## Implied intervals and imputation

`implied_interval()` reports the interval a cell could occupy for the matrix to
be valid, given the others. Point it at a **missing** (`NA`) entry and it imputes
the interval that entry must have occupied — a forensic "impute and check":

```{r}
R_miss <- round(rbind(cbind(Sig, r), c(r, 1)), 2)
R_miss[1, 4] <- R_miss[4, 1] <- NA        # pretend this correlation was unreported
implied_interval(R_miss)
```

## Rounding rules, mixed precision, and missing cells

The box must match how the numbers were actually produced. If a table was
**truncated** rather than rounded, the error is one-sided and up to `10^(-d)`;
a mismatched symmetric box can exclude the true value. Pass the rule:

```{r}
check_corr_psd(R_ok, decimals = 2, rounding = "truncate")$verdict
```

Mixed-precision tables (say, one variable reported at 1 decimal) take a
`decimals` matrix:

```{r}
dec <- matrix(2L, 3, 3); dec[1, ] <- dec[, 1] <- 1L
check_corr_psd(R_ok, decimals = dec)$verdict
```

A symmetric `NA` cell is freed to `[-1, 1]`, so the verdict holds for *every*
value the missing entry could take. Freeing a cell can rescue a matrix — or
not, which is then the stronger claim "inconsistent whatever the unreported value
was":

```{r}
R_na <- R_bad; R_na[1, 2] <- R_na[2, 1] <- NA
check_corr_psd(R_na, decimals = 2)$verdict     # some value of (1,2) would work

blk <- matrix(c(1, 0.9, 0.9, 0.9, 1, -0.9, 0.9, -0.9, 1), 3, 3)
J <- matrix(0, 6, 6); J[1:3, 1:3] <- blk; J[4:6, 4:6] <- blk; diag(J) <- 1
J[2, 3] <- J[3, 2] <- NA                        # free a cell of triangle 1
check_corr_psd(J, decimals = 2)$verdict         # triangle 2 still inconsistent
```

## Disattenuated correlations

Reported correlations are often corrected for measurement (un)reliability —
Spearman's disattenuation, `D_ij = R_ij / sqrt(rho_i · rho_j)`. Because
reliabilities are ≤ 1 this *inflates* every correlation, so a valid observed
matrix can become an **inconsistent** construct matrix under the reliabilities the
authors themselves report. `check_disattenuated_psd()` checks that, taking the
rounding of both the correlations and the reliabilities into account.

With a low reliability, the correction easily pushes correlations past 1:

```{r}
R_obs <- matrix(c(1, 0.60, 0.55, 0.60, 1, 0.50, 0.55, 0.50, 1), 3, 3)
check_disattenuated_psd(R_obs, reliability = 0.34, decimals = 2, max_plausible_r = 0.9)
```

`max_plausible_r` adds a third verdict level between `consistent` and `inconsistent`:
a corrected correlation above the cutoff (but ≤ 1) is `consistent but implausible` rather than
outright inconsistent — two supposedly distinct constructs correlating at 0.97 is a
red flag, not a mathematical contradiction:

```{r}
check_disattenuated_psd(R_obs, reliability = 0.62, decimals = 2, max_plausible_r = 0.9)
```

The inconsistency can also bind on positive semidefiniteness rather than on any
single correlation exceeding 1 — the subtler failure only this tool catches.

When reliabilities are **not** reported, pass `reliability = NULL` to get the
*critical reliability*: how reliable the measures would have had to be for the
disattenuation to be valid. An optional `plausible_floor` turns that into a
reachability statement against what the measures could realistically attain:

```{r}
check_disattenuated_psd(R_obs, reliability = NULL, decimals = 2,
                        max_plausible_r = 0.9, plausible_floor = 0.3)
```

The critical reliability is governed by the observed matrix's smallest
eigenvalue: `rho*_psd = 1 − lambda_min(R)`, with a second constraint
`rho*_range = max|R_ij|` (below which a corrected correlation exceeds 1). The
inconsistency side reuses the same witness/Rump engine as `check_corr_psd()`, so
it carries the same floating-point guarantee.

## A note on the engine

Everything above uses one self-contained numerical engine: a verified
witness-vector bound and a one-sided Cholesky test (Rump 2006) for the rigorous
sides, projections onto convex sets (POCS) as the search, and interval arithmetic
for the triple scan. There is no dependency on an external SDP solver. See the
package README for the full method provenance and references.
