---
title: "Getting started with psdness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with psdness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

```{r setup}
library(psdness)
```

## The question

A correlation matrix is only valid if it is **positive semidefinite** (PSD).
Reported matrices in papers, though, are *rounded* — you see `0.62`, not the
real number behind it. So the honest question is not "is this exact matrix PSD?"
but

> given that every off-diagonal was rounded to `d` decimals, does **any** PSD
> matrix exist inside the rounding box around the reported values?

`psdness` answers that **box** question. Writing `delta = 0.5 * 10^(-d)`, the box
is `X_ii = 1`, and for `i != j`, `X_ij` in
`[R_ij - delta, R_ij + delta]` intersected with `[-1, 1]`. The verdict is one of:

- **`impossible`** — no PSD matrix fits the box, so the reported matrix cannot be
  a genuine correlation matrix (a rigorous certificate; this is the payload);
- **`possible`** — a PSD matrix in the box is exhibited (or none was ruled out);
- **`undecided`** — precision-limited; the box grazes the PSD boundary.

## Checking a matrix

A genuine correlation matrix, rounded, comes back `possible`:

```{r}
R_ok <- matrix(c(1,    0.30, 0.20,
                 0.30, 1,    0.25,
                 0.20, 0.25, 1), 3, 3)
check_corr_psd(R_ok, decimals = 2)
```

An inconsistent one comes back `impossible`. The classic case is a "frustrated
triangle": variables 1 and 2 are both strongly correlated with 3, yet reported as
strongly *negatively* correlated with each other.

```{r}
R_bad <- matrix(c(1,    0.9,  0.9,
                  0.9,  1,   -0.9,
                  0.9, -0.9,  1), 3, 3)
res <- check_corr_psd(R_bad, decimals = 2)
res
```

### The certificate and the rigor guarantee

An `impossible` verdict carries a **witness vector** `v`: a direction along which
no in-box matrix has a nonnegative quadratic form, so none can be PSD. Extract it
with `certificate()`:

```{r}
certificate(res)
```

The guarantee is that the verdict is **floating-point-safe**: the witness
quantity `v'Rv + delta * (||v||_1^2 - ||v||_2^2)` is evaluated with a rigorous
a-priori error bound (Higham 2002), so `impossible` never rests on a solver
tolerance. A corollary is soundness in the other direction: because the bound is
a sound over-estimate, a box that contains any PSD matrix can never be reported
`impossible`. In particular a real correlation matrix rounded to `d` decimals is
never a false positive.

## Screening a corpus

`check_corr_psd_batch()` runs the check over a list of matrices and returns a
tidy table — useful for screening many reported matrices at once.

```{r}
mats <- list(ok = R_ok, frustrated_triangle = R_bad, identity = diag(4))
check_corr_psd_batch(mats, decimals = 2, quiet = TRUE)
```

## Localizing the fault

When a matrix is impossible, `localize_psd_fault()` asks **where** the problem is
and **how severe** it is. Non-PSDness is a global property, so a unique culprit
is often under-identified; the tool reports an honest localization class rather
than inventing one. The verdict is one of `cell`, `cell_tentative`, `triad`,
`variable`, `joint`, `diffuse`, or `none`.

### A frustrated triangle is a `triad`

Fixing *any one* of the three correlations would resolve the violation, so the
fault is not separable to a single cell:

```{r}
localize_psd_fault(R_bad, decimals = 2)
```

### A single bad cell

Here one transcription-style error makes a 5-variable matrix impossible. The tool
localizes it to the offending cell, states the interval that cell *should* have
occupied, and how far beyond rounding it must move:

```{r}
R_cell <- matrix(c( 1,    0.56,  0.68,  0.05,  0.82,
                    0.56, 1,    -0.26, -0.31, -0.75,
                    0.68, -0.26, 1,    -0.18,  0.50,
                    0.05, -0.31, -0.18, 1,     0.27,
                    0.82, -0.75, 0.50,  0.27,  1), 5, 5, byrow = TRUE)
localize_psd_fault(R_cell, decimals = 2)
```

### A joint fault (no single cell suffices)

Two independent inconsistencies mean no single edit repairs the matrix:

```{r}
blk <- matrix(c(1, 0.9, 0.9, 0.9, 1, -0.9, 0.9, -0.9, 1), 3, 3)
R_joint <- matrix(0, 6, 6); R_joint[1:3, 1:3] <- blk; R_joint[4:6, 4:6] <- blk
diag(R_joint) <- 1
localize_psd_fault(R_joint, decimals = 2)$localization_verdict
```

### The raw evidence

Every verdict carries the full box-aware evidence behind it — per-cell implied
intervals (A), rounding-robust impossible triples (B), leave-one-out (C), the
sparse minimal correction (D), and the R² / VIF variable localizer. Get it with
`fault_evidence()`:

```{r}
ev <- fault_evidence(localize_psd_fault(R_bad, decimals = 2))
names(ev)
```

The impossible-triple scan (component B) is also exposed on its own; it is pure
interval arithmetic and needs no solver:

```{r}
impossible_triples(R_bad, decimals = 2)
```

## Benign causes come first

Non-PSDness is *often benign*. A composite score plus its subscores, or a full
set of category dummies, is exactly rank-deficient by construction — not a sign
of a data problem (van Tilburg & van Tilburg 2023; Lorenzo-Seva & Ferrando
2021). For a forensic tool this matters, so every impossible result carries a
`structural` diagnosis that separates a **benign near-boundary** rank-deficiency
from a **substantive over-the-boundary** violation.

Here a composite equal to the sum of three subscores is nudged just past the
rounding boundary. The cellular verdict is `diffuse` (there is no single cell to
blame), but the structural read names the likely cause:

```{r}
Sig <- matrix(0.5, 3, 3); diag(Sig) <- 1
a <- c(1, 1, 1); r <- as.numeric(Sig %*% a) / sqrt(sum(a * (Sig %*% a)))
R_comp <- rbind(cbind(Sig, r), c(r, 1))
for (k in 1:3) R_comp[k, 4] <- R_comp[4, k] <- r[k] + 0.02   # nudge past the boundary
lf <- localize_psd_fault(round(R_comp, 2), decimals = 2)
lf$structural$severity_class
lf$structural$pattern
```

Contrast this with the frustrated triangle, whose severity is far beyond rounding:

```{r}
localize_psd_fault(R_bad, decimals = 2)$structural$severity_class
```

## Plausibility, even when possible

When a matrix passes the impossibility test it can still sit implausibly close to
the boundary. For `possible` matrices the result reports a **plausibility
gradient**: the per-variable reported-point R² and its closest approach to the
100% ceiling. The exactly-singular composite matrix (before the nudge) rounds to
`possible`, but one variable's R² exceeds 100% at face value:

```{r}
R_exact <- round(rbind(cbind(Sig, r), c(r, 1)), 2)   # composite = sum of subscores
lf0 <- localize_psd_fault(R_exact, decimals = 2)
lf0$localization_verdict        # "none": possible given rounding
lf0$plausibility$max_r2         # ... but a variable's reported-point R^2 exceeds 1
```

## Implied intervals and imputation

`implied_interval()` reports the interval a cell could occupy for the matrix to
be valid, given the others. Point it at a **missing** (`NA`) entry and it imputes
the interval that entry must have occupied — a forensic "impute and check":

```{r}
R_miss <- round(rbind(cbind(Sig, r), c(r, 1)), 2)
R_miss[1, 4] <- R_miss[4, 1] <- NA        # pretend this correlation was unreported
implied_interval(R_miss)
```

## A note on the engine

Everything above uses one self-contained numerical engine: a verified
witness-vector bound and a one-sided Cholesky test (Rump 2006) for the rigorous
sides, projections onto convex sets (POCS) as the search, and interval arithmetic
for the triple scan. There is no dependency on an external SDP solver. See the
package README for the full method provenance and references.
