---
title: "multiclosure walk-through"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multiclosure walk-through}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4
)
```

# Forensic reconstruction of a (synthetic) psychology Table 1

This vignette walks through `multiclosure` on a synthetic but realistic
psychology Table 1. The scenario: a hypothetical 6-item Likert (1–7) scale
"ResilienceX" administered to N = 217 participants. The Table reports per-item
means, SDs, a correlation matrix, and Cronbach's α, all rounded to 2dp.

```{r}
library(multiclosure)
dat <- readRDS(system.file("extdata", "synthetic_table1.rds",
                           package = "multiclosure"))
str(dat, max.level = 1L)
```

## Stage 1 — cheap impossibility certificates

```{r}
tols <- make_tols()
s1 <- run_stage1(N = dat$N, K = dat$K, M = dat$M, s = dat$s, R = dat$R,
                 alpha = dat$alpha, tols = tols)
s1$passed
s1$details$grim_grimmer
s1$details$bhatia_davis
s1$details$alpha
head(s1$details$frechet)
```

## Stage 2 — per-item CLOSURE enumeration

```{r}
marg <- enumerate_marginals(dat$M, dat$s, dat$N, dat$K)
sapply(marg, nrow)   # per-item candidate counts
```

## Stage 3 — beta-binomial plausibility filter

```{r}
fits <- lapply(seq_along(dat$M),
               function(j) bb_mom_fit(dat$M[j], dat$s[j], dat$K, dat$N))
sapply(fits, function(f) c(alpha = f$alpha, beta = f$beta,
                            skew_bb = f$skew_bb))

filtered <- lapply(seq_along(dat$M),
                   function(j) bb_skew_filter(marg[[j]], fits[[j]], dat$N))
sapply(filtered, function(f) f$retention)
```

## Stages 4 + 5 end-to-end

```{r}
out <- multi_closure(
  N = dat$N, K = dat$K, M = dat$M, s = dat$s, R = dat$R, alpha = dat$alpha,
  max_candidates = 50L,
  stage4 = list(strategy = "bb_closest", max_combos = 200L,
                samples_per_combo = 1L),
  verbose = FALSE, seed = 42L
)
out
```

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(out, what = "retention")
plot(out, what = "skew_bb")
```

## Adversarial variants

**Fréchet-violating correlation.** Bump one off-diagonal to 0.95:

```{r}
R_bad <- dat$R
R_bad[1, 2] <- R_bad[2, 1] <- 0.95
out_fr <- multi_closure(N = dat$N, K = dat$K, M = dat$M, s = dat$s,
                        R = R_bad, alpha = dat$alpha,
                        verbose = FALSE)
out_fr$feasibility
```

**Cronbach α inconsistency.** Keep R and s consistent with α ≈ 0.82 but report
0.30:

```{r}
out_a <- multi_closure(N = dat$N, K = dat$K, M = dat$M, s = dat$s,
                       R = dat$R, alpha = 0.30, verbose = FALSE)
out_a$feasibility
```

## Interpretation

* `feasibility$passed = TRUE` means at least one integer dataset on `{1..K}`
  reproduces the reported summary to within ±0.005 per scalar.
* `feasibility$passed = FALSE` with a Stage-1 reason means the summary is
  internally inconsistent — no dataset can produce it.
* `feasibility$passed = NA` means Stage 1 passed but Stage 4 NORTA could not
  find a verified candidate. A non-Gaussian-copula reconstruction may still
  exist; Stage 6 (MIP fallback) is needed and is planned for v2.
* Retention < 0.10 on any item means the reported (mean, SD) for that item is
  jointly atypical for a bounded-support beta-binomial generating process —
  consider floor/ceiling effects, contamination, or mixture composition.
