---
title: "Checking the metaBUS correlation matrices with `port`"
subtitle: "A forensic screen of 12,547 reported correlation matrices"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: show
    theme: cosmo
    fig-width: 7
    fig-height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 110)

library(port)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(furrr)

theme_set(theme_minimal(base_size = 12))

# Parallel backend for the batched deep dive below, matching the validation
# simulations' furrr setup.
plan(multisession)

`%||%` <- function(a, b) if (is.null(a)) b else a   # port's helper is internal

# Small on-disk cache so the expensive screens run once and re-renders are fast.
# The .rds caches live under data/processed/; they are derived and can be deleted.
cache <- function(file, expr) {
  if (file.exists(file)) return(readRDS(file))
  val <- force(expr)
  saveRDS(val, file)
  val
}
```

## What this is

[metaBUS](http://metabus.org) is a curated database of correlations harvested from the applied-psychology literature. The object `matrixList` holds **`r format(12547L, big.mark = ",")` reported correlation matrices**, one per source, with the variables named and every off-diagonal rounded to the two decimals that are the field's reporting convention.

A correlation matrix is a genuine (complete-data, single-sample) Pearson matrix only if it is **positive semidefinite** (PSD). But the numbers we see are *rounded*, so the honest question is the **box** question that `port` answers:

> given that every off-diagonal was rounded to two decimals, does **any** PSD matrix exist inside the rounding box around the reported values?

`port` returns one of three verdicts per matrix:

- **`inconsistent`** — no PSD matrix fits the rounding box (a rigorous,
  floating-point-safe certificate: the reported values cannot be a
  complete-data Pearson matrix, up to rounding);
- **`consistent`** — a PSD matrix inside the box is exhibited;
- **`undecided`** — precision-limited; the box grazes the PSD boundary.

A caveat for *this* corpus, stated up front: metaBUS matrices are often **meta-analytically assembled** — different cells come from different studies and samples. Such synthetic matrices are *routinely* non-PSD for entirely benign reasons (as are pairwise-deletion, polychoric, and disattenuated matrices). So the headline is **not** the count of inconsistent matrices, but their **severity**: how far past the PSD boundary they sit, which is what separates a benign near-boundary rounding/assembly artefact from a substantive error.

```{r load-data}
load("../data/raw/matrixList.RData")   # provides `matrixList`

p <- vapply(matrixList, nrow, integer(1))

tibble(matrices = length(matrixList),
       variables_min = min(p), 
       variables_median = median(p), 
       variables_max = max(p)) |>
  kable(caption = "The corpus")
```

### The precision assumption

Nearly all cells are reported to two decimals, so we screen at `decimals = 2` (`delta = 0.005`). This is **sound** for any cell reported to two *or more* decimals: assuming two decimals gives a box at least as wide as the true rounding, so a genuinely valid matrix is never falsely flagged. The only cells it could mishandle are ones truly reported at *one* decimal (a handful of matrices); those we would, if anything, under-flag, never over-flag.

```{r precision}
dp <- function(x) { x <- abs(x[!is.na(x)]); k <- 0L
  while (k < 6L && any(abs(x * 10^k - round(x * 10^k)) > 1e-9)) k <- k + 1L; k }

maxdp <- vapply(matrixList, function(M) dp(M[upper.tri(M)]), integer(1))

tibble(`finest decimal place used` = maxdp) |>
  count(`finest decimal place used`) |>
  mutate(`% of corpus` = round(100 * n / sum(n), 1)) |>
  kable(caption = "Reported precision (max decimal places per matrix)")
```

## Screening the whole corpus

`check_corr_psd_batch()` runs the rigorous check over every matrix and returns a tidy table (`id`, `verdict`, `certified`, `margin`, `delta`, `p`, ...). This is the one step that touches all `r format(length(matrixList), big.mark = ",")` matrices; it takes a couple of minutes and is cached.

```{r screen}
screen <- cache("../data/processed/metabus_screen.rds",
                check_corr_psd_batch(matrixList, decimals = 2, quiet = TRUE))

screen |>
  count(verdict) |>
  mutate(`%` = round(100 * n / sum(n), 1)) |>
  arrange(desc(n)) |>
  kable(caption = "Verdict distribution across the corpus")
```

```{r screen-plot, echo=FALSE}
screen |>
  count(verdict) |>
  mutate(verdict = reorder(verdict, n)) |>
  ggplot(aes(verdict, n, fill = verdict)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = format(n, big.mark = ",")), hjust = -0.1, size = 3.5) +
  coord_flip(clip = "off") +
  scale_fill_manual(values = c(inconsistent = "#c0392b", consistent = "#27ae60",
                               undecided = "#f39c12", error = "grey50")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "matrices", title = "port verdicts on the metaBUS corpus")
```

Inconsistency should track matrix size — more variables means more triples that can frustrate one another:

```{r by-size, echo=FALSE}
screen |>
  mutate(p_bin = cut(p, breaks = c(2, 4, 6, 8, 10, 15, 20, 30, Inf),
                     labels = c("3-4", "5-6", "7-8", "9-10", "11-15", "16-20", "21-30", "31+"))) |>
  group_by(p_bin) |>
  summarise(n = n(), `% inconsistent` = round(100 * mean(verdict == "inconsistent"), 1)) |>
  kable(caption = "Inconsistent share by matrix size")
```

## How clear-cut are the inconsistencies?

Every inconsistent verdict from the witness tier carries a **margin** — the value `B = v'Rv + delta*(||v||_1^2 - ||v||_2^2)` of the certificate direction `v`. It is already in the screen table, at no extra cost, and its sign is the certificate: `B < 0` proves no in-box matrix is PSD. The *more* negative it is, the further the box sits inside the infeasible region, so its distribution gives a corpus-wide read on how marginal the violations are (it is a certificate value, not a per-cell size, so read it qualitatively rather than comparing exact magnitudes across matrices of different `p`). The distribution is not concentrated at zero: a minority of matrices sit within a decimal place of consistency, but most lie clearly past the boundary, with a heavy tail of deeply inconsistent tables.

```{r margin, echo=FALSE}
incon <- filter(screen, verdict == "inconsistent", is.finite(margin), margin < 0)

ggplot(incon, aes(-margin)) +
  geom_histogram(bins = 50, fill = "#c0392b", colour = "white", linewidth = 0.1) +
  scale_x_continuous(trans = "log10") +
  labs(x = "-margin  (certificate depth, log scale)", y = "matrices",
       title = "Inconsistency runs from shallow to deep",
       subtitle = sprintf("%d witness-certified inconsistent matrices (median -margin %.2f)",
                          nrow(incon), median(-incon$margin)))
```

## Correlation distributions and predictors of inconsistency

Because `check_corr_psd_batch()` now reports the per-matrix off-diagonal correlation summary (`r_min`, `r_max`, `r_mean`, `r_sd`) alongside the number of variables `p`, we can describe the corpus and ask which features of a table predict inconsistency — all from the screen table, at no extra cost.

```{r corr-features}
feat <- screen |>
  filter(verdict != "error") |>
  mutate(max_abs_r = pmax(abs(r_min), abs(r_max)),
         inconsistent = verdict == "inconsistent")

feat |>
  transmute(`min r` = r_min, `max r` = r_max, `mean r` = r_mean,
            `SD of r` = r_sd, `max |r|` = max_abs_r, `n variables` = p) |>
  pivot_longer(everything(), names_to = "quantity", values_to = "v") |>
  group_by(quantity) |>
  summarise(mean = round(mean(v), 3), sd = round(sd(v), 3),
            min = round(min(v), 3), median = round(median(v), 3),
            max = round(max(v), 3), .groups = "drop") |>
  kable(caption = "Per-matrix correlation summaries across the corpus")
```

### Correlation structure vs matrix size

Larger tables tend to have a slightly tighter spread of correlations, but the key point is how size and spread jointly relate to inconsistency (below).

```{r corr-vs-p, echo=FALSE}
feat |>
  group_by(p) |>
  summarise(mean_r_sd = mean(r_sd), mean_max_abs = mean(max_abs_r), n = n(),
            .groups = "drop") |>
  ggplot(aes(p, mean_r_sd)) +
  geom_line(colour = "grey70") +
  geom_point(aes(size = n), alpha = 0.6, colour = "#2c3e50") +
  scale_size_area(max_size = 5) +
  labs(x = "number of variables (p)", y = "mean SD of off-diagonal correlations",
       size = "matrices", title = "Correlation spread vs matrix size")
```

### What distinguishes the inconsistent tables

```{r corr-by-verdict, echo=FALSE}
feat |>
  filter(verdict %in% c("consistent", "inconsistent")) |>
  select(verdict, `mean r` = r_mean, `SD of r` = r_sd, `max |r|` = max_abs_r,
         `n variables` = p) |>
  pivot_longer(-verdict, names_to = "quantity", values_to = "v") |>
  ggplot(aes(v, fill = verdict)) +
  geom_density(alpha = 0.5, colour = NA) +
  facet_wrap(~ quantity, scales = "free") +
  scale_fill_manual(values = c(consistent = "#27ae60", inconsistent = "#c0392b")) +
  labs(x = NULL, y = "density", fill = NULL,
       title = "Consistent vs inconsistent tables")
```

Inconsistency rate rises with both the spread of correlations and their largest magnitude:

```{r rate-by-bins, echo=FALSE}
rate_by <- function(var) {
  br <- unique(quantile(feat[[var]], seq(0, 1, 0.2), na.rm = TRUE))
  feat |>
    mutate(bin = cut(.data[[var]], br, include.lowest = TRUE)) |>
    group_by(bin) |>
    summarise(n = n(), `% inconsistent` = round(100 * mean(inconsistent), 1),
              .groups = "drop")
}
rate_by("r_sd")      |> kable(caption = "Inconsistency rate by SD-of-r quintile")
rate_by("max_abs_r") |> kable(caption = "Inconsistency rate by max |r| quintile")
```

### A joint model

A logistic regression on standardized predictors puts the effects on one scale (odds ratio per SD of each feature):

```{r logit}
md <- feat |>
  filter(verdict %in% c("consistent", "inconsistent")) |>
  transmute(y        = as.integer(inconsistent),
            z_mean   = as.numeric(scale(r_mean)),
            z_sd     = as.numeric(scale(r_sd)),
            z_maxabs = as.numeric(scale(max_abs_r)),
            z_p      = as.numeric(scale(p)))
fit <- glm(y ~ z_mean + z_sd + z_maxabs + z_p, data = md, family = binomial)
co  <- summary(fit)$coefficients
tibble(term = c("(intercept)", "mean r", "SD of r", "max |r|", "n variables"),
       `odds ratio / SD` = round(exp(co[, 1]), 2),
       z = round(co[, 3], 1),
       `p value` = signif(co[, 4], 2)) |>
  kable(caption = "Logistic regression: standardized predictors of inconsistency")
```

The table quantifies which features carry the most weight. Geometrically, more variables and larger correlation magnitudes create more opportunities for a frustrated triple, so both should push the odds of inconsistency up; the mean correlation level matters less than the spread and the extremes.

## Severity and localization (resumable, batched deep dive)

The interpretable severity number is `excusable_delta()`: the smallest per-cell reporting imprecision that would make a matrix consistent, in the same units as `delta`. And `localize_psd_fault()` says *where* the violation lives and whether it is structurally benign. Both are far heavier than the screen (seconds per matrix), so the deep dive processes the inconsistent matrices **in parallel batches (via `furrr`/`future`), writing each completed batch to disk**. Each batch is analysed across worker processes; a run then resumes from what is already saved and works only on the matrices still outstanding, so the full set can be completed across several renders — or survive an interruption — without repeating any work. `MAX_PER_RUN` bounds how many *new* matrices a single render processes (the full set is large; at a few seconds each a complete pass takes a while, so cap it and re-render to continue).

```{r deep-setup}
DEEP_FILE   <- "../data/processed/metabus_deep.rds"
BATCH_SIZE  <- 25L      # matrices processed before each write to disk
MAX_PER_RUN <- Inf      # cap on NEW matrices per render (e.g. 200L to bound a run)

dir.create(dirname(DEEP_FILE), showWarnings = FALSE, recursive = TRUE)

# Target set: every inconsistent matrix (swap in a sample here if desired).
target_ids <- screen$id[screen$verdict == "inconsistent"]

# One pass per matrix: severity, localization, structural read, smallest
# eigenvalue, and the sole-culprit cells (their reported value and the interval
# they must occupy) captured as a nested tibble for the cell-level export below.
# Takes the matrix explicitly so only a batch's matrices ship to the workers.
analyse_one <- function(id, M) {
  lf   <- localize_psd_fault(M, decimals = 2)
  sev  <- lf$severity
  st   <- lf$structural
  recs <- lf$evidence$sole_culprit_cells               # component A: sole culprits
  pull <- function(field) if (length(recs))
    vapply(recs, function(c) as.numeric(c[[field]]), numeric(1)) else numeric(0)
  sole_cells <- tibble(
    i = if (length(recs)) vapply(recs, function(c) as.integer(c$i), integer(1)) else integer(0),
    j = if (length(recs)) vapply(recs, function(c) as.integer(c$j), integer(1)) else integer(0),
    reported = pull("reported"),
    implied_lo = pull("lo"), implied_hi = pull("hi"),
    required_edit = pull("required_edit"))
  imp <- lf$implicated$cells
  tibble(
    id = id, p = nrow(M),
    min_eig = min(eigen(M, symmetric = TRUE, only.values = TRUE)$values),
    margin = sev$witness_margin,
    excusable_delta = sev$excusable_delta,
    severity_max = sev$severity_max,
    severity_frob = sev$severity_frob,
    best_lambda_min = sev$best_lambda_min,
    localization_verdict = lf$localization_verdict,
    severity_class = if (is.null(st$severity_class)) NA_character_ else st$severity_class,
    severity_ratio = if (is.null(st$severity_ratio)) NA_real_ else st$severity_ratio,
    n_sole = nrow(sole_cells),
    implicated_cells = if (is.null(imp)) NA_character_ else
      paste(sprintf("(%d,%d)", imp[, 1], imp[, 2]), collapse = "; "),
    sole_cells = list(sole_cells))
}

# furrr options: parallel-safe RNG stream + load port/tibble on the workers.
fo <- furrr_options(seed = TRUE, packages = c("port", "tibble"))

empty_deep <- tibble(
  id = character(), p = integer(), min_eig = double(), margin = double(),
  excusable_delta = double(), severity_max = double(), severity_frob = double(),
  best_lambda_min = double(), localization_verdict = character(),
  severity_class = character(), severity_ratio = double(),
  n_sole = integer(), implicated_cells = character(), sole_cells = list())
```

```{r deep-run}
# Resume from whatever has already been completed on disk.
deep <- if (file.exists(DEEP_FILE)) readRDS(DEEP_FILE) else empty_deep

# If the cache predates the enriched schema, it cannot supply the cell-level
# detail, so rebuild it from scratch.
if (!all(names(empty_deep) %in% names(deep))) {
  message("deep-dive cache has an older schema; rebuilding from scratch.")
  deep <- empty_deep
}

todo <- utils::head(setdiff(target_ids, deep$id), MAX_PER_RUN)

if (length(todo) > 0L) {
  batches <- split(todo, ceiling(seq_along(todo) / BATCH_SIZE))
  for (b in batches) {
    # each batch is analysed in parallel across workers (furrr), then the
    # completed batch is written to disk before the next one starts.
    rows <- future_pmap(list(id = b, M = matrixList[b]), analyse_one, .options = fo)
    deep <- bind_rows(deep, bind_rows(rows))
    saveRDS(deep, DEEP_FILE)                       # persist the completed batch
    message(sprintf("deep dive: %d / %d inconsistent matrices done",
                    sum(deep$id %in% target_ids), length(target_ids)))
  }
}

# Keep only rows for the current target set (in case the file holds extras).
deep <- deep[deep$id %in% target_ids, , drop = FALSE]

n_done <- nrow(deep); n_target <- length(target_ids)
cat(sprintf("Deep dive: %d of %d inconsistent matrices completed (%.0f%%)%s.\n",
            n_done, n_target, 100 * n_done / n_target,
            if (n_done < n_target) " -- re-render to continue" else ""))
```

### Severity in precision units

```{r severity-band}
deep |>
  mutate(band = case_when(
    excusable_delta <= 0.05 ~ "<= 0.05  (excusable by 1-decimal reporting)",
    excusable_delta <= 0.10 ~ "0.05-0.10  (mild)",
    excusable_delta <= 0.20 ~ "0.10-0.20  (moderate)",
    TRUE                    ~ "> 0.20  (substantive: far beyond rounding)")) |>
  count(band) |>
  mutate(`%` = round(100 * n / sum(n), 1)) |>
  kable(caption = "Severity of the completed inconsistent matrices, in precision units")
```

```{r severity-plot, echo=FALSE}
ggplot(deep, aes(excusable_delta)) +
  geom_histogram(bins = 25, fill = "#c0392b", colour = "white", linewidth = 0.1) +
  geom_vline(xintercept = c(0.05, 0.20), linetype = 2, colour = "grey40") +
  scale_x_continuous(trans = "log10") +
  labs(x = "excusable_delta (log scale)", y = "matrices",
       title = "Smallest imprecision that would excuse each matrix",
       subtitle = "Dashed: 1-decimal (0.05) and substantive (0.20) marks")
```

### Where the fault sits, and whether it is benign

```{r localize-class}
deep |>
  count(localization_verdict) |>
  mutate(`%` = round(100 * n / sum(n), 1)) |>
  arrange(desc(n)) |>
  kable(caption = "Localization classes (completed)")

deep |>
  count(severity_class) |>
  mutate(`%` = round(100 * n / sum(n), 1)) |>
  kable(caption = "Structural severity class (completed)")
```

## Export for human inspection

To support manual checking of the worst-offending articles (e.g. for data-extraction errors), the completed deep dive is written to three CSVs under `data/processed/`:

- **`metabus_offenders.csv`** — one row per inconsistent matrix, worst first: severity (`excusable_delta`, `severity_max`, smallest eigenvalue), localization class, structural read, the implicated cells, and the correlation summary.
- **`metabus_offender_cells.csv`** — cell-level detail for every *sole-culprit* cell (a single cell whose value alone makes the table inconsistent): the reported correlation, its rounding box, and the interval it would have to occupy for the table to be valid, with the `required_edit` beyond rounding. A reported value far outside its implied interval is the classic signature of a transcription/extraction error.
- **`metabus_offender_matrices_top.csv`** — the full off-diagonal of the `TOP_N` most severe matrices in long form with variable names, so a reviewer can reconstruct each table and compare it to the source.

```{r export, message=FALSE}
OUT_DIR <- dirname(DEEP_FILE)
DELTA   <- 0.005                          # nearest-rounding half-width at 2 dp

# 1. Offender triage table (one row per inconsistent matrix, worst first).
offenders <- deep |>
  select(-sole_cells) |>
  left_join(select(screen, id, r_min, r_max, r_mean, r_sd), by = "id") |>
  arrange(desc(excusable_delta)) |>
  transmute(
    id, n_variables = p,
    min_eigenvalue = round(min_eig, 4),
    certificate_margin = round(margin, 4),
    excusable_delta = round(excusable_delta, 4),
    severity_max = round(severity_max, 4),
    severity_frob = round(severity_frob, 4),
    localization = localization_verdict,
    severity_class,
    severity_ratio = round(severity_ratio, 1),
    n_sole_culprit_cells = n_sole,
    implicated_cells,
    r_min = round(r_min, 3), r_max = round(r_max, 3),
    r_mean = round(r_mean, 3), r_sd = round(r_sd, 3))
write.csv(offenders, file.path(OUT_DIR, "metabus_offenders.csv"), row.names = FALSE)

# 2. Cell-level detail for the sole-culprit cells: reported value vs the
#    interval it must occupy, with the rounding box and the required edit.
add_cell_context <- function(id, tbl) {
  if (is.null(tbl) || nrow(tbl) == 0L) return(NULL)
  M  <- matrixList[[id]]
  vn <- rownames(M); if (is.null(vn)) vn <- as.character(seq_len(nrow(M)))
  tibble(
    id = id, i = tbl$i, j = tbl$j, var_i = vn[tbl$i], var_j = vn[tbl$j],
    reported_r = tbl$reported,
    rounding_lo = round(tbl$reported - DELTA, 4),
    rounding_hi = round(tbl$reported + DELTA, 4),
    implied_lo = round(tbl$implied_lo, 4),
    implied_hi = round(tbl$implied_hi, 4),
    required_edit = round(tbl$required_edit, 4))
}
cells <- bind_rows(Map(add_cell_context, deep$id, deep$sole_cells))
if (nrow(cells) > 0L) {
  cells <- cells |>
    left_join(select(offenders, id, localization, severity_class, excusable_delta),
              by = "id") |>
    arrange(desc(excusable_delta), id, desc(abs(required_edit)))
}
write.csv(cells, file.path(OUT_DIR, "metabus_offender_cells.csv"), row.names = FALSE)

# 3. Full matrices of the worst offenders (long format, with variable names).
TOP_N <- 100L
worst_ids <- head(offenders$id, TOP_N)
mat_long <- bind_rows(lapply(worst_ids, function(id) {
  M  <- matrixList[[id]]
  vn <- rownames(M); if (is.null(vn)) vn <- as.character(seq_len(nrow(M)))
  ut <- which(upper.tri(M), arr.ind = TRUE)
  tibble(id = id, i = ut[, 1], j = ut[, 2],
         var_i = vn[ut[, 1]], var_j = vn[ut[, 2]], r = M[ut])
}))
write.csv(mat_long, file.path(OUT_DIR, "metabus_offender_matrices_top.csv"),
          row.names = FALSE)

cat(sprintf("Wrote to %s:\n  metabus_offenders.csv              (%d rows)\n  metabus_offender_cells.csv         (%d sole-culprit cells)\n  metabus_offender_matrices_top.csv  (%d cells from %d worst matrices)\n",
            OUT_DIR, nrow(offenders), nrow(cells), nrow(mat_long), length(worst_ids)))
```

```{r export-preview, echo=FALSE}
offenders |>
  head(10) |>
  select(id, n_variables, excusable_delta, severity_max, localization,
         severity_class, implicated_cells) |>
  kable(caption = "Worst 10 offenders (full detail in metabus_offenders.csv)")

if (nrow(cells) > 0L) {
  cells |>
    head(12) |>
    select(id, var_i, var_j, reported_r, implied_lo, implied_hi, required_edit) |>
    kable(caption = "Sole-culprit cells: reported value vs the interval it must occupy")
}
```

## Worked examples

Concrete matrices make the range tangible. We restrict the worked examples to the more **compact** completed matrices (<= 10 variables) so the certificates print legibly, and pull the **most severe** and a **marginal** inconsistency from them.

```{r examples-pick}
compact <- deep |> filter(p <= 10) |> arrange(desc(excusable_delta))

if (nrow(compact) < 2) compact <- arrange(deep, desc(excusable_delta))

worst_id    <- compact$id[1]
marginal_id <- compact$id[nrow(compact)]
```

### The most severe inconsistency (compact matrices)

```{r ex-worst}
worst <- matrixList[[worst_id]]
cat("source id:", worst_id, "  variables:", nrow(worst), "\n")
check_corr_psd(worst, decimals = 2)
```

```{r ex-worst-localize}
localize_psd_fault(worst, decimals = 2)
```

### A marginal inconsistency

Just past the boundary — the kind of thing rounding of an assembled table readily produces:

```{r ex-marginal}
marginal <- matrixList[[marginal_id]]
cat("source id:", marginal_id, "  variables:", nrow(marginal),
    "  excusable_delta:", round(compact$excusable_delta[nrow(compact)], 4), "\n")
check_corr_psd(marginal, decimals = 2)
```

### A consistent matrix

For contrast, a matrix the screen clears:

```{r ex-ok}
ok_id <- screen$id[screen$verdict == "consistent"][1]

check_corr_psd(matrixList[[ok_id]], decimals = 2)
```

## Summary

```{r summary, echo=FALSE}
n_incon <- sum(screen$verdict == "inconsistent")

tibble(
  metric = c("matrices screened",
             "inconsistent",
             "consistent",
             "undecided",
             "deep dive completed",
             "completed inconsistencies excusable by 1-decimal reporting (<= 0.05)"),
  value = c(format(nrow(screen), big.mark = ","),
            sprintf("%s (%.1f%%)", format(n_incon, big.mark = ","),
                    100 * n_incon / nrow(screen)),
            sprintf("%.1f%%", 100 * mean(screen$verdict == "consistent")),
            sprintf("%.1f%%", 100 * mean(screen$verdict == "undecided")),
            sprintf("%s of %s", format(nrow(deep), big.mark = ","),
                    format(n_incon, big.mark = ",")),
            if (nrow(deep) > 0L)
              sprintf("%.1f%%", 100 * mean(deep$excusable_delta <= 0.05)) else "n/a")
) |>
  kable(caption = "Screen summary")
```

**Reading it.** 

A non-trivial share (~6.5%) of metaBUS matrices are not positive semidefinite even allowing for rounding — unsurprising for an assembled corpus, where cells come from different samples. But the violations are *graded*, not uniform: a minority sit within about one decimal place of consistency (the benign signature of assembly and rounding), while the majority lie clearly past the boundary and a heavy tail is inconsistent by an amount no rounding can explain. `port` turns the blunt non-PSD fact into that gradient, and for each flagged matrix localizes the offending cells, states the interval they should have occupied, and certifies the verdict without any solver tolerance — so a reviewer can triage from the deep tail down.

The screen assumed two-decimal reporting; matrices in the substantive tail are worth re-checking against their original sources for transcription error, pairwise deletion, or disattenuation before inferring a mistake.

## Session info

```{r}
sessionInfo()
```
