---
title: "port validation — Sim 2: soundness under rounding"
author: "Ian Hussey"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc_float: true
    code-fold: false
    code-overflow: wrap
    code-tools: true
    warning: false
    df-print: kable
    theme:
      light: cosmo
      dark: darkly
    highlight-style:
      light: breeze
      dark: dracula
---

```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(scipen = 999)
```

# Introduction

This is the second of three validation simulations for `port`, following the ADEMP framework (Morris et al., 2019; Siepe et al., 2024). Sim 1 verified that the tool never certifies an exactly-valid matrix as inconsistent. Sim 2 verifies the same soundness property under **rounding**, and validates the boundary condition on it: the tool's assumed rounding tolerance (`delta`) must match the rounding rule that actually produced the reported values.

## Project description

A reported correlation matrix is almost always rounded. The package builds a box of half-width `delta` around the reported entries and asks whether any PSD matrix lies inside. The forensic soundness requirement is that rounding a genuine correlation matrix must never yield `"inconsistent"`.

## Prior related work

As in Sim 1, the mathematics is classical; what is validated is the implementation and, newly here, the interaction between the *rounding rule* and the tool's `delta` assumption.

## Aims

1. Verify the **false-inconsistency rate is 0** when a valid matrix is rounded to nearest with the matching symmetric `delta = 0.5·10^(-d)` (the theorem below).
2. Show that an **asymmetric rounding rule with a mismatched (too-small) `delta`** can make the tool report `"inconsistent"` — correctly for the box it was given, but misleadingly relative to the truth — and that a **matched `delta`** restores the zero-false-inconsistency property.

### Theorem (the pass condition)

For round-to-nearest with `delta = 0.5·10^(-d)`, every entry moves by at most `delta`, so the original PSD matrix `R` lies inside the box around the reported `R̃`. The box therefore always contains a PSD matrix and the tool must not return `"inconsistent"`. Any such verdict is a bug and fails loudly. For truncation / floor / ceiling the error can reach `10^(-d)`, so with the nearest-rule `delta` the true `R` may fall *outside* the box; supplying the matched `delta = 10^(-d)` puts `R` back inside and restores soundness.

## Data-Generating Process

### DGP specification approach and factors

Valid matrices are generated exactly as in Sim 1 (`onion` at eta ∈ {1, 0.5, 0.1}, and the `factor_model`), then **contaminated by rounding**. Factors: generator, dimension `p`, decimals `d`, `rounding_method` ∈ {nearest, truncate, floor, ceiling}, and a `delta_match` flag controlling whether the tool's `delta` matches the rule.

### Factor values and settings

- generator ∈ {onion(eta=1), onion(eta=0.5), onion(eta=0.1), factor_model}. The near-singular onion(eta=0.1) is the condition where a mismatched `delta` can actually bite.
- `p` ∈ {3, 5, 7, 10, 15, 20}; `d` ∈ {1, 2, 3}.
- `rounding_method` × `delta_match`: nearest (matched only; its natural `delta` is symmetric), and truncate/floor/ceiling each with unmatched (`0.5·10^(-d)`) and matched (`10^(-d)`).

### Factor combination and number of conditions

Fully factorial over the crossed factors, with 7 `(method, match)` combinations, giving `4 × 6 × 3 × 7 = 504` conditions, each replicated `K` times. Base matrices are reused across all conditions sharing `(generator, p, iteration)`.

## Estimands and Targets

The **target is the true feasibility status**, which is fixed by the theorem: the *matched* box always contains the original PSD `R`, so the truth is **feasible** in every replication. The performance measure is the empirical `"inconsistent"` rate; its target is 0 under matched delta, and a mismatched delta is expected to inflate it (a correct-but-misleading verdict about a mis-specified box).

## Methods and extracted quantities

`check_corr_psd(R̃, delta = delta_used)`. Extracted: `verdict`, `tier`. As a cross-check the independent oracle records the box status **at the tool's own delta**; where the tool says `"inconsistent"` under a mismatched delta, the oracle should confirm that box is genuinely infeasible — i.e. the tool is sound about the box it was handed, and the harm is a delta-specification error, not a tool bug.

## Performance and Uncertainty

- **Primary — false-inconsistency rate** `P(verdict == "inconsistent")`, target 0 (against the matched-box truth), split by `rounding_method`, `delta_match`, generator, `p`, `d`. Wilson 95% intervals (the estimate hugs 0 in the sound conditions).
- **Secondary — completeness**: consistent vs undecided by generator/p/decimals.
- **Hard assertion**: any `"inconsistent"` in a **matched-delta** condition is a soundness bug (fail loudly, log seed + matrix).

`K = 1000` baseline (`K = 10000` for a publishable zero-rate claim; `SIM_QUICK=1` for development).

# Methods

## Dependencies

```{r}
library(tidyr)
library(dplyr)
library(readr)
library(furrr)
library(janitor)
library(simhelpers)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)

library(port)
source("R/sim_helpers.R")

plan(multisession)
```

## Functions

Shared with the other sims via `R/sim_helpers.R`: `generate_data()`, `contaminate_data()` (rounds off-diagonals only, mirrors, keeps the diagonal at 1), `analyse_check()`, and the independent `box_is_feasible()` oracle. The per-step simulation functions are defined here and called by name in the run below.

```{r}
# generate a valid base matrix (with generator-appropriate defaults)
gen_base <- function(generator, p, seed, eta, loading_sd) {
  generate_data(generator, p, seed,
                eta = if (is.na(eta)) 1 else eta,
                loading_sd = if (is.na(loading_sd)) 0.35 else loading_sd)
}

# contaminate: round the off-diagonals under the chosen rule
contaminate_round <- function(R, decimals, rounding_method) {
  contaminate_data(R, decimals, rounding_method)$R_tilde
}

# run the checker at the tool's (possibly mismatched) delta
check_delta <- function(R_tilde, delta) analyse_check(R_tilde, delta = delta)

# independent oracle status at the tool's own delta
oracle_delta <- function(R_tilde, delta, R) {
  tibble::tibble(oracle_status = box_is_feasible(R_tilde, delta, witness = R)$status)
}
```

## Define experiment

```{r}
K <- if (Sys.getenv("SIM_QUICK") == "1") 40L else 1000L

generator_grid <- tibble::tribble(
  ~generator,     ~gen_label,        ~eta, ~loading_sd,
  "onion",        "onion(eta=1)",       1,          NA,
  "onion",        "onion(eta=0.5)",   0.5,          NA,
  "onion",        "onion(eta=0.1)",   0.1,          NA,
  "factor_model", "factor_model",      NA,        0.50
)

# rounding rule x tool-delta combinations. delta_mult multiplies 10^(-d):
# 0.5 = the nearest-rule (symmetric) delta; 1.0 = the matched delta for an
# asymmetric rule. nearest appears once (matched == its natural delta).
method_grid <- tibble::tribble(
  ~rounding_method, ~delta_match, ~delta_mult,
  "nearest",        "matched",           0.5,
  "truncate",       "unmatched",         0.5,
  "truncate",       "matched",           1.0,
  "floor",          "unmatched",         0.5,
  "floor",          "matched",           1.0,
  "ceiling",        "unmatched",         0.5,
  "ceiling",        "matched",           1.0
)

# base matrices: one per (generator, p, iteration), reused across conditions
base_grid <- expand_grid(
  gen_label = generator_grid$gen_label,
  p         = c(3, 5, 7, 10, 15, 20),
  iteration = seq_len(K)
) |>
  left_join(generator_grid, by = "gen_label") |>
  mutate(seed = as.integer((GENERATOR_INDEX[generator] * 1e7 + p * 1e4 + iteration) %%
                             .Machine$integer.max))

experiment_parameters_grid <- base_grid |>
  cross_join(expand_grid(decimals = c(1, 2, 3),
                         method_id = seq_len(nrow(method_grid)))) |>
  left_join(mutate(method_grid, method_id = row_number()), by = "method_id") |>
  mutate(delta_used = delta_mult * 10^(-decimals),
         condition = paste0(gen_label, ", p=", p, ", d=", decimals,
                            ", ", rounding_method, "/", delta_match))
```

## Run simulation

```{r}
if(file.exists("simulation2.rds")){
  simulation <- read_rds("simulation2.rds")
} else {
  set.seed(42)
  
  fo <- furrr_options(seed = TRUE, packages = c("port", "clusterGeneration"))
  
  # Step 1: generate the base matrix (list column, reused via the shared seed).
  base <- base_grid |>
    mutate(R = future_pmap(
      .l = list(generator = generator, p = p, seed = seed, eta = eta, loading_sd = loading_sd),
      .f = gen_base, .options = fo, .progress = TRUE))

  simulation <- experiment_parameters_grid |>
    left_join(select(base, gen_label, p, iteration, R), by = c("gen_label", "p", "iteration")) |>
    # Step 2: contaminate (round off-diagonals under the chosen rule)
    mutate(R_tilde = future_pmap(
      list(R = R, decimals = decimals, rounding_method = rounding_method),
      .f = contaminate_round, .options = fo, .progress = TRUE)) |>
    # Step 3: analyse with the tool's delta
    mutate(check = future_pmap(list(R_tilde = R_tilde, delta = delta_used),
                               .f = check_delta, .options = fo, .progress = TRUE)) |>
    # Step 4: independent oracle at the tool's own delta (sound-about-its-box check)
    mutate(oracle = future_pmap(list(R_tilde = R_tilde, delta = delta_used, R = R),
                                .f = oracle_delta, .options = fo)) |>
    select(-R, -R_tilde)
  
  write_rds(simulation, "simulation2.rds", compress = "gz")
}
```

# Results

## Soundness diagnostics (hard failure on any matched-delta false-inconsistent)

```{r}
checked <- simulation |> unnest(check) |> unnest(oracle)

matched_violations <- checked |>
  filter(delta_match == "matched", verdict == "inconsistent")

if (nrow(matched_violations) > 0) {
  matched_violations |>
    select(gen_label, p, decimals, rounding_method, iteration, seed, verdict) |>
    kable(caption = "SOUNDNESS VIOLATIONS under matched delta") |>
    kable_styling(full_width = FALSE)
  stop("SOUNDNESS VIOLATION: a valid matrix rounded with a MATCHED delta was ",
       "certified inconsistent. Offending seeds listed above.")
} else {
  cat("PASS: 0 false-inconsistent verdicts across",
      sum(checked$delta_match == "matched"), "matched-delta (sound) conditions.\n")
}
```

The unmatched-delta inconsistents are expected. This chunk confirms that where the tool *does* say inconsistent under a mismatched delta, the oracle agrees the box (at that delta) is genuinely infeasible — i.e. the tool is sound about the box it was given:

```{r}
checked |>
  filter(delta_match == "unmatched", verdict == "inconsistent") |>
  count(oracle_confirms_infeasible = oracle_status == "infeasible") |>
  kable(caption = "Unmatched-delta inconsistents: does the oracle confirm the (mis-specified) box is infeasible?") |>
  kable_styling(full_width = FALSE)
```

## Primary: false-inconsistency rate by rounding rule and delta match

```{r}
sim2_summary <- checked |>
  group_by(rounding_method, delta_match, generator = gen_label) |>
  summarise(K = n(),
            n_inconsistent = sum(verdict == "inconsistent"),
            false_inconsistent_rate = mean(verdict == "inconsistent"),
            prop_consistent = mean(verdict == "consistent"),
            prop_undecided = mean(verdict == "undecided"),
            .groups = "drop") |>
  rowwise() |>
  mutate(wilson_lo = wilson_ci(n_inconsistent, K)["lower"],
         wilson_hi = wilson_ci(n_inconsistent, K)["upper"]) |>
  ungroup()
```

```{r}
sim2_summary |>
  group_by(rounding_method, delta_match) |>
  summarise(n_inconsistent = sum(n_inconsistent), K = sum(K),
            false_inconsistent_rate = n_inconsistent / K, .groups = "drop") |>
  rowwise() |>
  mutate(ci = paste0("[", scales::number(wilson_ci(n_inconsistent, K)["lower"], accuracy = 0.0001),
                     ", ", scales::number(wilson_ci(n_inconsistent, K)["upper"], accuracy = 0.0001), "]")) |>
  ungroup() |>
  mutate(rate = scales::number(false_inconsistent_rate, accuracy = 0.0001)) |>
  select(rounding_method, delta_match, false_inconsistent_rate = rate, wilson_ci = ci, n_inconsistent, K) |>
  kable(caption = "False-inconsistency rate by rounding rule x delta match (target 0; unmatched asymmetric expected > 0)") |>
  kable_styling(full_width = FALSE)
```

```{r}
sim2_summary |>
  group_by(rounding_method, delta_match) |>
  summarise(false_inconsistent_rate = sum(n_inconsistent) / sum(K), .groups = "drop") |>
  ggplot(aes(x = rounding_method, y = false_inconsistent_rate, fill = delta_match)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  scale_y_continuous(name = "False-inconsistency rate", labels = number_format(accuracy = 0.001)) +
  labs(x = "Rounding rule", fill = "Tool delta",
       title = "Mismatched delta on an asymmetric rule can fire; matched delta restores soundness") +
  theme_linedraw() + theme(panel.grid.minor = element_blank())
```

## Secondary: completeness (consistent vs undecided)

```{r}
checked |>
  filter(delta_match == "matched") |>
  group_by(gen_label, decimals) |>
  summarise(`% consistent` = scales::percent(mean(verdict == "consistent"), accuracy = 0.1),
            `% undecided` = scales::percent(mean(verdict == "undecided"), accuracy = 0.1),
            .groups = "drop") |>
  kable(caption = "Completeness under matched delta, by generator and decimals") |>
  kable_styling(full_width = FALSE)
```

# Discussion

## Performance of the method across conditions

Under nearest rounding with the matching symmetric `delta` — and under any matched-delta condition — the false-inconsistency rate is exactly 0, as the theorem requires; the diagnostics chunk would have halted the render otherwise. This confirms that the package's `delta` assumption, when correct, guarantees soundness under rounding.

Under an **asymmetric** rule (truncate/floor/ceiling) with the **nearest-rule delta**, the tool can report `"inconsistent"` — most visibly for the near-singular `onion(eta=0.1)` generator, where the box is already tight. These are not tool bugs: the second diagnostics table shows the oracle agrees that the box *at the mis-specified delta* is genuinely infeasible. The verdict is correct about the box it was handed; the error is that the box was too small for the rounding rule. Supplying the matched `delta = 10^(-d)` returns the rate to 0. The practical lesson is that `check_corr_psd()` must be told a `delta` consistent with how the reported numbers were produced.

## Conclusions with regard to the aims

Both aims are met: matched-delta soundness is exact (Aim 1), and the mismatch phenomenon plus its matched-delta fix are demonstrated (Aim 2).

## Limitations and intended use

- The mismatch effect is **rare** and concentrated at the near-singular boundary; a larger `K` is needed to estimate its magnitude precisely.
- Truncation toward zero tends to *increase* feasibility (smaller off-diagonals), so it rarely fires; floor/ceiling are the informative asymmetric rules here.
- `K = 1000`; publishable claims need `K = 10000`. The cross-join over conditions holds base matrices by shared reference, but the rounded-matrix column scales with the condition count — reduce the grid or `K` if memory-bound.

# References

Morris, T. P., White, I. R., & Crowther, M. J. (2019). Using simulation studies to evaluate statistical methods. *Statistics in Medicine*, 38(11), 2074–2102.

Siepe, B. S., Bartoš, F., Morris, T. P., Boulesteix, A.-L., Heck, D. W., & Pawel, S. (2024). Simulation studies for methodological research in psychology: A standardized template for planning, preregistration, and reporting. *Psychological Methods*.

# Session info

```{r}
sessionInfo()
```
